---
title: "Function Design Best Practices"
format: html
---

## Why Function Design Matters

Good function design reduces cognitive load for you and your users. The [Tidyverse design guide](https://design.tidyverse.org/) offers a few core principles:

- Programming is a task performed by humans
- Reduce cognitive load with consistent design
- Make your functions and systems composable
- Think about others who are not like us

Today we will refactor our `fishr` functions with these principles in mind.


## Naming Conventions

Some general guidelines:

- Use verbs to ascribe an action
- Use consistent style (e.g., `snake_case`)
- Consider short prefixes to unify package functions
- Don't be afraid to be verbose
- Avoid conflict with existing functions

```{r}
# bad
gse(date = "1977-05-25")

# good
get_salmon_escapement(date = "1977-05-25")
```

### Naming in fishr

Our functions follow these conventions:

- `cpue()` -- short but domain-standard abbreviation
- `biomass_index()` -- descriptive, easy to predict


## Pure Functions vs Side Effects

A **pure function** produces the same output for the same input and has no impact on anything outside itself. Pure functions are easier to test and reason about.

A function with **side effects** interacts with the outside environment (writes files, prints messages, modifies global state).

### Annotating our cpue function

Let's look at `cpue()` as it stands at the end of Day 1:

```{r}
#| eval: true
cpue <- function(catch, effort, gear_factor = 1, verbose = FALSE) {
  # Side effect: prints a message
  if (verbose) {
    message("Processing ", length(catch), " records")
  }

  # Pure calculation
  raw_cpue <- catch / effort

  raw_cpue * gear_factor
}
```

The calculation itself (`catch / effort * gear_factor`) is pure. The `message()` call is a side effect -- it interacts with the outside environment.

### Package-wide options

Sometimes it is helpful to turn off these side effects. R has a global key-value store called `options()` that packages can use for user-configurable defaults. You interact with it using two functions:

```{r}
# Set an option
options(fishr.verbose = TRUE)

# Get an option (with a fallback default)
getOption("fishr.verbose", default = FALSE)
```

By convention, package options use the package name as a prefix (`fishr.verbose`, not just `verbose`) to avoid confusion with other packages.

We can use `getOption()` directly in the function signature. This makes the option visible to the user (they'll see it in `?cpue`) and keeps the function body cleaner:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  verbose = getOption("fishr.verbose", default = FALSE)
) {
  if (verbose) {
    message("Processing ", length(catch), " records")
  }

  raw_cpue <- catch / effort

  raw_cpue * gear_factor
}
```

Now users can set `options(fishr.verbose = TRUE)` once and any `fishr` function that uses this option will pick it up, no need to pass `verbose = TRUE` every time. But users can still override it per-call:

```r
cpue(1, 5, verbose = TRUE)
```


### Bad side-effect examples

Avoid functions that silently affect the user's environment:

```{r}
# DON'T do this -- writes a file without asking
bad_cpue <- function(catch, effort) {
  result <- catch / effort
  write.csv(data.frame(cpue = result), "cpue_log.csv")
  result
}

# DON'T do this -- changes global options
bad_summary <- function(x) {
  options(digits = 2)
  summary(x)
}
```

The first function leaves behind a file the user likely didn't ask for, and is doing two things at once - writing a file _and_ returning the result. Functions that are called for their side-effects are ok (eg. `write.csv()`, `plot()`, etc) but should _only_ be called for their side effect. They should _usually_ return their input value. Sometimes they return a path (eg. if writing a file), or `NULL`.

The second changes how *all* numbers display for the rest of the session. Unlike our `fishr.verbose` option, which lets users opt *in* to behaviour, this forces a change on them. Return values instead and let the caller decide what to do with them.

### Our biomass_index function

```{r}
#| eval: true
biomass_index <- function(cpue, area_swept) {
  cpue * area_swept
}
```

This is a pure function -- no side effects at all. And it reliably returns the same output for the same set of inputs.


## Default Arguments and match.arg

### Adding a method parameter to cpue

Right now `cpue()` always calculates a simple ratio. Let's add a `method` argument so users can choose between ratio CPUE and log-transformed CPUE.

Update `R/cpue.R`:

```{r}
#' Calculate Catch Per Unit Effort (CPUE)
#'
#' Calculates CPUE from catch and effort data, with optional gear
#' standardization. Supports ratio and log-transformed methods.
#'
#' @param catch Numeric vector of catch (e.g., kg)
#' @param effort Numeric vector of effort (e.g., hours)
#' @param gear_factor Numeric scalar for gear standardization (default 1)
#' @param method Character; one of `"ratio"` (default) or `"log"`.
#' @param verbose Logical; print processing info? Default from
#'   `getOption("fishr.verbose", FALSE)`.
#'
#' @return A numeric vector of CPUE values
#' @export
#'
#' @examples
#' cpue(100, 10)
#' cpue(c(100, 200), c(10, 20), method = "log")
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = catch / effort,
    log = log(catch / effort)
  )

  raw_cpue * gear_factor
}
```

The `switch()` function selects which expression to evaluate based on the value of `method`. It's a clean alternative to a chain of `if`/`else if` statements when dispatching on a single value.

### How match.arg works

`match.arg()` takes the first element of the default vector when the user doesn't supply a value, and gives clear error messages for invalid input:

```{r}
load_all()

# Uses default ("ratio")
cpue(100, 10)

# Explicit
cpue(100, 10, method = "log")

# Invalid input gives a helpful error
cpue(100, 10, method = "median")
```

### Document and check

```{r}
document()

check()
```

::: {.callout-warning}
## Snapshot update needed

Adding `method` to the verbose message changed its format - from `"Processing 2 records"` to `"Processing 2 records using ratio method"`. If you have a snapshot test for the verbose output, `check()` will fail with a snapshot mismatch. Accept the updated snapshot before continuing:

```{r}
testthat::snapshot_accept()
```
:::

::: {.callout-tip}
## Make a commit
:::


## The Ellipsis (...)

### Passing arguments through with ...

The ellipsis `...` lets a function accept extra arguments and pass them to another function. This is useful when one function wraps another.

Let's refactor `biomass_index()` so it can optionally compute CPUE on the fly by accepting `catch` and `effort` instead of a pre-computed `cpue` value.

Update `R/biomass.R`:

```{r}
#' Calculate Biomass Index
#'
#' Calculates biomass index from CPUE and area swept. Can optionally
#' compute CPUE from catch and effort data.
#'
#' @param cpue Numeric vector of CPUE values. If `catch` and `effort` are
#'   provided, this is computed automatically.
#' @param area_swept Numeric vector of area swept (e.g., km²)
#' @param catch Optional numeric vector of catch. If provided with `effort`,
#'   CPUE is computed via `cpue()`.
#' @param effort Optional numeric vector of effort. Required if `catch` is
#'   provided.
#' @param ... Additional arguments passed to `cpue()` when computing from
#'   catch and effort (e.g., `method`, `gear_factor`).
#'
#' @return A numeric vector of biomass index values
#' @export
#'
#' @examples
#' # From pre-computed CPUE
#' biomass_index(cpue = 10, area_swept = 5)
#'
#' # Compute CPUE on the fly
#' biomass_index(area_swept = 5, catch = 100, effort = 10)
#'
#' # Pass method through to cpue()
#' biomass_index(
#'   area_swept = 5,
#'   catch = c(100, 200),
#'   effort = c(10, 20),
#'   method = "log"
#' )
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  if (is.null(cpue) && (!is.null(catch) && !is.null(effort))) {
    cpue <- cpue(catch, effort, ...)
  }

  if (is.null(cpue)) {
    stop("Must provide either 'cpue' or both 'catch' and 'effort'.")
  }

  cpue * area_swept
}
```

### Demo

```{r}
load_all()

# Pre-computed CPUE
biomass_index(cpue = 10, area_swept = 5)

# Compute on the fly
biomass_index(area_swept = 5, catch = 100, effort = 10)

# Pass method through to cpue()
biomass_index(
  area_swept = 5,
  catch = c(100, 200),
  effort = c(10, 20),
  method = "log"
)
```

Notice that in the second call, we wrote `area_swept = 5` even though it's the second argument. Because we're skipping `cpue`, positional matching would put `5` into `cpue` instead. When you skip optional arguments or call them out of order, you need to name them.

This is also why optional arguments default to `NULL` rather than a meaningful value. `cpue = NULL` signals "not provided" clearly, and `is.null()` is a reliable way to check. If we had used `cpue = 0` as the default, there would be no way to tell whether the user passed zero intentionally or just didn't provide a value.

### Pitfalls of ...

The ellipsis is powerful but has risks:

- Arguments passed through `...` must be named -- positional matching doesn't work through the ellipsis, so `biomass_index(area_swept = 5, catch = 100, effort = 10, "log")` won't do what you expect. Always use `method = "log"`.
- Misspelled arguments are silently absorbed -- `cpue(100, 10, mthod = "log")` won't error, it just ignores `mthod`
- Confusing error origins -- errors from inner functions can be hard to trace
- Use `rlang::check_dots_used()` at the top of your function to catch unused dots (covered below)

### Document and check

```{r}
document()

check()
```

::: {.callout-tip}
## Make a commit
:::


## Managing Dependencies

Every function your package calls from another package must be declared as a dependency.
Undeclared dependencies may work on your machine because the package happens to be installed, declaring them ensures they are available for users of your package. `R CMD check` will catch them.

### Declaring a dependency

Use `usethis::use_package()` to add a package to the `Imports` field of `DESCRIPTION`:

```{r}
use_package("rlang")
```

Before this call, `DESCRIPTION` has no `Imports` field. After:

```default
Imports:
    rlang
```

Now call functions from the package using `::` notation:

```{r}
format_catch <- function(catch) {
  if (!is.numeric(catch)) {
    rlang::abort("`catch` must be a numeric vector.")
  }
  round(catch, 2)
}
```

The `::` operator is the preferred approach for package development: it is explicit, avoids
namespace conflicts, and makes clear which package each function comes from.

### Importing frequently-used functions

For functions you call throughout a package, repeating `rlang::` on every call is verbose.
`usethis::use_import_from()` adds the function to your `NAMESPACE` and lets you call it directly:

```{r}
use_import_from("rlang", "abort")
```

This inserts a `@importFrom` tag into a roxygen block (usually `R/<package>-package.R`):

```r
#' @importFrom rlang abort
```

After running `document()`, `abort()` is available across your package without the `rlang::`
prefix. The `NAMESPACE` file is updated automatically - never edit it by hand.

### Catching unused dots

Now that `rlang` is declared, use `rlang::check_dots_used()` to catch misspelled arguments
before they are silently absorbed:

```{r}
# Example: catching unused dots
f <- function(x, ...) {
  rlang::check_dots_used()
  mean(x, ...)
}

f(1:10) # works
f(1:10, na.rm = TRUE) # works
f(1:10, narm = TRUE) # errors: unused dots
```

Apply this to `biomass_index()`:

```{r}
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  rlang::check_dots_used()

  if (is.null(cpue) && (!is.null(catch) && !is.null(effort))) {
    cpue <- cpue(catch, effort, ...)
  }

  if (is.null(cpue)) {
    stop("Must provide either 'cpue' or both 'catch' and 'effort'.")
  }

  cpue * area_swept
}
```

### Document and check

```{r}
document()

check()
```

::: {.callout-tip}
## Make a commit
:::


## DRY: Don't Repeat Yourself

### Why validate inputs?

Our functions currently accept any input -- what happens if someone passes a character?

```{r}
cpue("one hundred", 10)
```

We get an unhelpful error (or worse, silent nonsense). Let's add validation. But rather than copy-paste the same check into every function, we'll create a reusable helper.

### Create a helper file

```{r}
use_r("utils")
```

### Write a validation helper

Add to `R/utils.R`:

```{r}
#' Validate that inputs are numeric
#'
#' Checks each named argument and stops with an informative error
#' if any are not numeric.
#'
#' @param ... Named numeric inputs to validate.
#'
#' @return Invisible `NULL`. Called for its side effect of
#'   stopping with an error if validation fails.
#'
#' @noRd
validate_numeric_inputs <- function(...) {
  args <- list(...)
  arg_names <- names(args)

  for (i in seq_along(args)) {
    if (!is.numeric(args[[i]])) {
      stop(
        "'",
        arg_names[i],
        "' must be numeric, got ",
        class(args[[i]])[1],
        ".",
        call. = FALSE
      )
    }
  }

  invisible(NULL)
}
```

The `@noRd` tag tells roxygen2 not to generate a `.Rd` file -- this is an internal helper, not part of the public API.

### Add validation to cpue

Update `R/cpue.R`:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)

  validate_numeric_inputs(catch = catch, effort = effort)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = catch / effort,
    log = log(catch / effort)
  )

  raw_cpue * gear_factor
}
```

### Add validation to biomass_index

Update `R/biomass.R`:

```{r}
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  rlang::check_dots_used()

  if (is.null(cpue) && (!is.null(catch) && !is.null(effort))) {
    cpue <- cpue(catch, effort, ...)
  }

  if (is.null(cpue)) {
    stop("Must provide either 'cpue' or both 'catch' and 'effort'.")
  }

  validate_numeric_inputs(cpue = cpue, area_swept = area_swept)

  cpue * area_swept
}
```

### Verify the helper works

```{r}
load_all()

# Good input
cpue(100, 10)

# Bad input -- now shows which argument is the problem
cpue("high", 10)

biomass_index(cpue = "ten", area_swept = 5)
```

### Where to put helper functions

When you extract a helper, you have two choices:

- Same file as the exported function -- when the helper is specific to that one function. For example, if `cpue()` needed a helper function to clean up effort values, put it below `cpue()` in `R/cpue.R`. This keeps related logic together and makes it easy to find later.
- A shared file like `R/utils.R` -- when the helper is used by multiple functions across the package. `validate_numeric_inputs()` is a good example: both `cpue()` and `biomass_index()` use it, so it belongs in its own file.

A good rule of thumb: start with the helper in the same file. Move it to `utils.R` only when a second function needs it.

### A note on function length

There's no hard rule, but if a function exceeds ~20-30 lines, consider whether it's doing too many things. Smaller functions are easier to test, read, and reuse.

### Document and check

```{r}
document()

check()
```

::: {.callout-tip}
## Make a commit
:::


## Function Composition

Good function design pays off when functions compose well together. Each function does one thing and can be combined with others.

### Composed workflow

```{r}
load_all()

# Step by step
my_cpue <- cpue(catch = c(100, 200, 300), effort = c(10, 20, 30))
biomass_index(cpue = my_cpue, area_swept = 50)

# Or in one call, thanks to ... pass-through
biomass_index(area_swept = 50, catch = c(100, 200, 300), effort = c(10, 20, 30))

# With options passed through
biomass_index(
  area_swept = 50,
  catch = c(100, 200, 300),
  effort = c(10, 20, 30),
  method = "log"
)
```

### Recap

Composability is the payoff of good function design:

- **Naming** -- clear names make composed code readable
- **Pure calculations** -- predictable functions are safe to chain
- **match.arg** -- consistent interfaces reduce mistakes
- **...** -- pass-through enables flexible composition
- **Helpers** -- shared validation keeps behaviour consistent

### Final check

```{r}
check()
```
