---
title: Object-Oriented Programming in R
---

## Introduction

You already use object-oriented programming (OOP) patterns constantly in R, even if you have not thought about it explicitly.

Let's look at the `plot()` and `summary()` functions.

```{r}
x_norm <- rnorm(5000)
plot(x_norm)

summary(mtcars)
plot(mtcars)
plot(mtcars$mpg, mtcars$hp)

mtcars_model <- lm(mpg ~ wt, data = mtcars)
summary(mtcars_model)
plot(mtcars_model)
```

The same function - `summary()` - does something completely different depending on what you
hand it. A numeric vector gets a five-number summary. A data frame gets column summaries. A linear model fit gets coefficients, standard errors,
and p-values. The function name stays the same; the behaviour
adapts to the data.

This is one of the most powerful ideas in OOP: you write code that calls `print()` or `plot()` or
`summary()`, and the object itself determines what happens.

We are going to:

1. Create a new S3 class `cpue_result` to wrap CPUE calculations with metadata
2. Write `print`, `summary`, and `plot` methods for it
3. Convert `cpue()` and `biomass_index()` into generics that work with both numeric vectors and data frames

## OOP Systems in R

R has several OOP systems. They differ in formality and where methods live.

| System | Dispatch | Methods live on... | Formality | Common in |
|--------|----------|-------------------|-----------|-----------|
| **S3** | Single (first argument) | The generic function | Informal | Base R, tidyverse, most packages |
| **S4** | Multiple arguments | The generic function | Formal (classes, validity) | Bioconductor, methods-heavy packages |
| **S7** | Multiple arguments | The generic function | Formal (classes, properties, validation) | Some new packages (successor to S3 and S4) |
| **R6** | Single | The object itself | Moderate (via R6 package) | Mutable state, encapsulation |

For package development, S3 covers the vast majority of use cases.
S4 is worth learning if you work in Bioconductor or need multiple dispatch.
S7 is designed as the eventual successor to both S3 and S4 - it combines the simplicity of S3 with more of the rigor of S4 (formal class definitions, property validation, multiple dispatch). 
If you are starting a new package and want more structure than S3 provides, S7 is worth considering, though S3 remains the dominant system in practice.

We will focus on S3 today, with a brief look at R6 at the end.

## S3 in Depth

S3 is the most widely used OOP system in R. It is simple, flexible, and what base R itself uses.

### How S3 works

An S3 object is just a regular R object (list, vector, data frame, etc.) with a `class` attribute.
When you call a generic function like `print(x)`, R looks at `class(x)` and dispatches to the
appropriate method.

```{r}
class(x_norm)
class(mtcars)
class(mtcars_model)
```

### Seeing the dispatch chain

```{r}
# What methods exist for print?
methods(generic.function = "print")

# What methods exist for a class?
methods(class = "data.frame")
```

### The dispatch mechanism

When you call `summary(x)` where `x` has class `"foo"`:

1. R looks for `summary.foo`
2. If not found, looks for `summary.default`
3. If neither exists, throws an error

This is called **single dispatch** - the method is chosen based on the class of _the first argument only_.

## Creating an S3 Class

Our `cpue()` function returns a plain numeric vector:

```{r}
result <- cpue(c(100, 200, 300), c(10, 20, 15))
class(result)
```

That class is just `"numeric"`. When we `print(result)`, R dispatches to `print.numeric` and
shows raw numbers. What if we want CPUE output to identify itself - to carry context about the
calculation and display it in a useful way?

### The simplest way: `class() <-`

The most direct way to create an S3 class is to set the `class` attribute on an existing object.
Let's modify `cpue()` to tag its output. Add two lines at the end of the function body in
`R/cpue.R`:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)

  validate_numeric_inputs(catch = catch, effort = effort)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = catch / effort,
    log = log(catch / effort)
  )

  result <- raw_cpue * gear_factor
  class(result) <- "cpue_result"
  result
}
```

```{r}
load_all()

result <- cpue(c(100, 200, 300), c(10, 20, 15))
result
class(result)
```

Now R knows this is a `cpue_result`. We can write a `print` method for it. 

## Writing Methods

### print method

The `print` method controls what users see when they type an object name at the console. 
When you type `result` and hit Enter, R calls `print(result)` behind the scenes - 
this is called **autoprinting**. R looks at the class of the object, finds `print.cpue_result`, and calls that. 
Without a custom print method, R falls back to `print.default`, which is why our 
tagged vector currently displays as raw numbers.

```{r}
print.cpue_result <- function(x, ...) {
  cat("CPUE Results for", length(x), "records\n")
  cat("Values:", round(x, 2), "\n")
  invisible(x)
}
```

Important details:

- The function name must be `print.<classname>`
- Always include `...` in the signature (required by the `print` generic)
- Return `invisible(x)` so assignment still works: `y <- print(x)`
- Use `cat()` rather than `print()` inside print methods to avoid recursion

```{r}
document()
load_all()

result
```


That is the entire mechanism behind S3 dispatch: set a class, write a `function.classname`
method, and R connects the two automatically.

Note that although `result` *looks* different when we print it, it is still a numeric vector under the hood:

```{r}
is.numeric(result)
unclass(result)
```

### Adding metadata with attributes

A bare numeric vector with a class label has no context. What gear factor was
used? What method was used? R objects can carry arbitrary **attributes** - named
metadata attached to the object. Let's update the function body again:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)

  validate_numeric_inputs(catch = catch, effort = effort)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = catch / effort,
    log = log(catch / effort)
  )

  result <- raw_cpue * gear_factor
  attr(result, "gear_factor") <- gear_factor
  attr(result, "n_records") <- length(catch)
  attr(result, "method") <- method
  class(result) <- "cpue_result"
  result
}
```

```{r}
load_all()

result <- cpue(c(100, 200, 300), c(10, 20, 15))
attr(result, "gear_factor")
attr(result, "n_records")
attr(result, "method")
```

Now our print method can be more informative:

```{r}
#' @export
print.cpue_result <- function(x, ...) {
  cat("Survey Result\n")
  cat("Records:", attr(x, "n_records"), "\n")
  cat("Gear factor:", attr(x, "gear_factor"), "\n")
  cat("Method:", attr(x, "method"), "\n")
  cat("CPUE values:", round(x, 2), "\n")
  invisible(x)
}
```

As a class accumulates more fields, setting attributes one by one gets unwieldy. There
is a better pattern.

### Create a constructor function

For anything beyond a very basic class, the standard practice is to create a **constructor function** that builds the object in one place.
A common naming convention, recommended in [Advanced R](https://adv-r.hadley.nz/s3.html#s3-constructor), is `new_<classname>`.

```{r}
#' Create a cpue result object
#'
#' @param cpue_values Numeric vector of CPUE values.
#' @param method Character string indicating the calculation method.
#' @param gear_factor Numeric gear correction factor used.
#' @param n_records Integer number of records processed.
#'
#' @return A `cpue_result` object.
#' @noRd
new_cpue_result <- function(cpue_values, method, gear_factor, n_records) {
  structure(
    cpue_values, # The data
    method = method, # Attributes specifying metadata
    gear_factor = gear_factor,
    n_records = n_records,
    class = "cpue_result" # class is a special attribute
  )
}
```

A few things changed:

- `structure()` sets multiple attributes at once - including `class` - instead of separate
  `attr()` and `class() <-` calls
- All object creation is centralised

```{r}
#' Summarize a CPUE survey
#'
#' Calculates CPUE from catch and effort data and returns a structured
#' `cpue_result` object with metadata about the calculation.
#'
#' @param catch Numeric vector of catch values.
#' @param effort Numeric vector of effort values.
#' @param gear_factor Numeric gear correction factor (default 1).
#' @param method Calculation method: "ratio" or "log".
#'
#' @return A `cpue_result` object.
#' @export
#'
#' @examples
#' cpue(catch = c(100, 200, 300), effort = c(10, 20, 30))
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  # ...cpue body here

  new_cpue_result(
    cpue_values = result,
    method = method,
    gear_factor = gear_factor,
    n_records = length(catch)
  )
}
```

```{r}
load_all()

result <- cpue(
  catch = c(100, 200, 300, 150),
  effort = c(10, 20, 15, 30)
)

result
```

### summary method

`summary` should compute and return a useful statistical summary:

```{r}
#' @export
summary.cpue_result <- function(object, ...) {
  stats <- list(
    method = attr(object, "method"),
    n_records = attr(object, "n_records"),
    gear_factor = attr(object, "gear_factor"),
    mean_cpue = mean(object),
    median_cpue = stats::median(object),
    sd_cpue = stats::sd(object),
    range_cpue = range(object)
  )

  structure(stats, class = "summary.cpue_result")
}

#' @export
print.summary.cpue_result <- function(x, ...) {
  cat("Survey Result Summary\n")
  cat("---------------------\n")
  cat("Method:      ", x$method, "\n")
  cat("Records:     ", x$n_records, "\n")
  cat("Gear factor: ", x$gear_factor, "\n")
  cat("Mean CPUE:   ", round(x$mean_cpue, 2), "\n")
  cat("Median CPUE: ", round(x$median_cpue, 2), "\n")
  cat("SD CPUE:     ", round(x$sd_cpue, 2), "\n")
  cat(
    "CPUE Range:    ",
    round(x$range_cpue[1], 2),
    "-",
    round(x$range_cpue[2], 2),
    "\n"
  )
  invisible(x)
}
```

Note that `summary` returns an object of class `summary.cpue_result`, which has its own print method.
This is the standard pattern in base R (see `summary.lm`).

```{r}
document()
load_all()

summary(result)
``` 

### Your turn: Make a `plot()` method

Plot methods must be named `plot.<classname>`, take `x` as the first argument, and include `...` in the signature.

Note that you don't need to use base `plot()` in a plot method - you can use ggplot2 or any other plotting system. 

```{r}
#' @export
plot.cpue_result <- function(x, ...) {
  plot(
    seq_along(x),
    x,
    type = "b",
    xlab = "Record",
    ylab = "CPUE",
    main = paste("CPUE -", attr(x, "method"), "method"),
    ...
  )
}
```

The `...` passes through to `plot.default`, allowing users to customize (colours, line types, etc.)
without us having to anticipate every option.

```{r}
document()
load_all()

plot(result)
plot(result, col = "steelblue", pch = 19)
```

### Registering S3 methods

For your methods to work when the package is installed (not just with `load_all()`), they need to
appear in the NAMESPACE file. The `@export` tag on each method handles this via roxygen2.

```{r}
document()
```

Look at the NAMESPACE file. What do you see different about S3 methods compared to regular functions?

```
S3method(print,cpue_result)
S3method(summary,cpue_result)
S3method(plot,cpue_result)
```

### Testing the class

We should update our tests to verify that `cpue()` returns the right class. testthat provides `expect_s3_class()` for exactly this:

```{r}
test_that("cpue() returns a cpue_result object", {
  result <- cpue(c(100, 200), c(10, 20))
  expect_s3_class(result, "cpue_result")
})
```

We can also test that the metadata attributes are set correctly:

```{r}
test_that("cpue_result carries calculation metadata", {
  result <- cpue(c(100, 200, 300), c(10, 20, 15), method = "log")
  expect_equal(attr(result, "method"), "log")
  expect_equal(attr(result, "gear_factor"), 1)
  expect_equal(attr(result, "n_records"), 3)
})
```

And snapshot tests are a natural fit for print methods, since they capture the exact console output:

```{r}
test_that("print.cpue_result displays expected output", {
  result <- cpue(c(100, 200, 300), c(10, 20, 15))
  expect_snapshot(print(result))
})
```

::: {.callout-tip}
## Make a commit
:::

## Creating your own generics and methods

So far we have built new classes and created methods for existing generics. But one of the most practical uses
of S3 is creating your own generics, so the same function name works
on different types of input.

Our `cpue()` function currently takes numeric vectors:

```{r}
cpue(catch = c(100, 200), effort = c(10, 20))
```

But users often have their data in a data frame. Instead of making them extract columns every
time, we can make `cpue()` work directly on data frames too. The same function call adapts its
behaviour depending on what the user passes in - just like `summary()` does in base R.

### Step 1: Make cpue a generic

We replace the function body with `UseMethod("cpue")`. The first argument becomes `x` by
convention, since it could be either a numeric vector or a data frame:

```{r}
#' Calculate Catch Per Unit Effort (CPUE)
#'
#' Calculates CPUE from catch and effort data, with optional gear
#' standardization. Supports ratio and log-transformed methods.
#'
#' @param x Input data: a numeric vector of catch values, or a data frame
#'   containing catch and effort columns.
#' @param ... Additional arguments passed to methods.
#'
#' @export
cpue <- function(x, ...) {
  UseMethod("cpue")
}
```

### Step 2: Move the existing implementation into cpue.numeric

The existing code becomes the numeric method. The first argument changes from `catch` to `x`
to match the generic's signature, but the logic is identical:

```{r}
#' @rdname cpue
#'
#' @param effort Numeric vector of effort (e.g., hours)
#' @param gear_factor Numeric scalar for gear standardization (default 1)
#' @param method Character; one of `"ratio"` (default) or `"log"`.
#' @param verbose Logical; print processing info? Default from
#'   `getOption("fishr.verbose", FALSE)`.
#'
#' @return A numeric vector of CPUE values
#' @export
#'
#' @examples
#' cpue(100, 10)
#' cpue(c(100, 200), c(10, 20), method = "log")
cpue.numeric <- function(
  x,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE),
  ...
) {
  method <- match.arg(method)

  validate_numeric_inputs(x = x, effort = effort)

  if (verbose) {
    message("Processing ", length(x), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = x / effort,
    log = log(x / effort)
  )

  new_cpue_result(
    cpue_values = raw_cpue * gear_factor,
    method = method,
    gear_factor = gear_factor,
    n_records = length(x)
  )
}
```

All existing code that calls `cpue()` with positional arguments continues to work - R
dispatches to `cpue.numeric` because the first argument is numeric.

::: {.callout-warning}
## Named argument change

The first argument was renamed from `catch` to `x` to match the generic's signature. Any call using named arguments - `cpue(catch = 100, effort = 10)` - will now fail because `catch` is no longer a parameter. Update these to use positional arguments: `cpue(100, 10)`.
:::

### Step 3: Add a data frame method

```{r}
#' @rdname cpue
#' @param catch_col Name of the catch column (default `"catch"`).
#' @param effort_col Name of the effort column (default `"effort"`).
#' @export
cpue.data.frame <- function(
  x,
  catch_col = "catch",
  effort_col = "effort",
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE),
  ...
) {
  if (!catch_col %in% names(x)) {
    stop("Column '", catch_col, "' not found in data frame.", call. = FALSE)
  }
  if (!effort_col %in% names(x)) {
    stop("Column '", effort_col, "' not found in data frame.", call. = FALSE)
  }

  # We can then call the numeric method by extracting the relevant columns and passing them to cpue() again.
  # This way we reuse the existing logic and maintain a single source of truth for the CPUE calculation.
  cpue(
    x = x[[catch_col]],
    effort = x[[effort_col]],
    gear_factor = gear_factor,
    method = method,
    verbose = verbose,
    ...
  )
}
```

The data frame method extracts the right columns and calls `cpue()` again with numeric vectors,
which dispatches to `cpue.numeric`. This layering - where one method calls the generic again
with different input - is a common and effective pattern.

### Step 4: Add a `.default` method

It's good practice to add a default method that throws an informative error if 
the user passes in an unsupported type. If the generic can't find a method for 
the class of `x`, it will fall back to `cpue.default`:


```{r}
#' @rdname cpue
#' @export
cpue.default <- function(x, ...) {
  stop("Unsupported input type for cpue(): ", class(x), call. = FALSE)
}
```

### Update biomass_index documentation

Converting `cpue` into a generic changed its signature from `cpue(catch, effort, ...)` to `cpue(x, ...)`. The `@inheritParams cpue` tag on `biomass_index` was inheriting `catch` and `effort` docs from the old function - those params no longer exist on the generic.

Update `biomass_index` to inherit from `cpue.numeric` instead, since that is the method it actually calls. Because `cpue.numeric` names its first argument `x` rather than `catch`, add `@param catch` back explicitly:

```{r}
#' Calculate Biomass Index
#'
#' @param cpue Numeric vector of CPUE values. If NULL, computed from `catch`
#'   and `effort`.
#' @param area_swept Numeric vector of area swept (e.g., km²).
#' @param catch Numeric vector of catch (e.g., kg).
#' @inheritParams cpue.numeric
#' @inheritDotParams cpue.numeric -effort
#' @export
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  # ...
}
```

```{r}
document()
load_all()

# Still works with vectors
cpue(c(100, 200), c(10, 20))

# Now also works with data frames
fishing_data <- data.frame(
  catch = c(100, 200, 300),
  effort = c(10, 20, 15)
)

cpue(fishing_data)

# Custom column names
trawl_data <- data.frame(
  kg = c(50, 100, 150),
  hours = c(5, 10, 8)
)

cpue(trawl_data, catch_col = "kg", effort_col = "hours")
```

### Updating and adding tests

Two things have changed that may break existing tests:

- **Named argument calls:** tests using `cpue(catch = ..., effort = ...)` will fail now that the argument is named `x`. Change these to positional calls.
- **Return type:** `cpue()` now returns a `cpue_result` object, not a plain numeric. `expect_equal()` compares attributes, so tests that compare the result to a bare number will fail. Wrap the result in `as.numeric()` for numeric comparisons: `expect_equal(as.numeric(cpue(100, 10)), 10)`.

Some existing snapshots will also be stale since the print format changed. Run `snapshot_accept()` after `check()` to accept the updated snapshots.

Now add tests for the data frame method and the default error:

```{r}
test_that("cpue.data.frame dispatches correctly", {
  fishing_data <- data.frame(
    catch = c(100, 200, 300),
    effort = c(10, 20, 15)
  )
  result <- cpue(fishing_data)
  expect_s3_class(result, "cpue_result")
  expect_equal(as.numeric(result), c(10, 10, 20))
})

test_that("cpue.data.frame works with custom column names", {
  trawl_data <- data.frame(
    kg = c(50, 100),
    hours = c(5, 10)
  )
  result <- cpue(trawl_data, catch_col = "kg", effort_col = "hours")
  expect_s3_class(result, "cpue_result")
  expect_equal(as.numeric(result), c(10, 10))
})

test_that("cpue.data.frame errors on missing columns", {
  df <- data.frame(x = 1, y = 2)
  expect_snapshot(cpue(df), error = TRUE)
})

test_that("cpue.default gives informative error", {
  expect_snapshot(cpue("not valid"), error = TRUE)
})
```

::: {.callout-tip}
## Make a commit
:::

<!--
TODO: This is still very awkward with the way biomass_index() is structured
where the first argument can be missing.

## Your Turn: biomass_index.data.frame

`biomass_index()` can accept either pre-computed CPUE values or raw catch/effort data (with `...` passing arguments through to `cpue()`). Convert it into a generic with a numeric method and a data frame method.

Hint: follow the same three steps we used for `cpue()`:

1. Replace the function body with `UseMethod("biomass_index")`
2. Move the existing implementation into `biomass_index.numeric` - keep the optional `catch`/`effort`/`...` pass-through
3. Write `biomass_index.data.frame` that extracts columns and delegates to the numeric method

::: {.callout-tip collapse="true"}

## Solution

```{r}
#' Calculate Biomass Index
#'
#' @param x Input data: a numeric vector of CPUE values, or a data frame
#'   containing CPUE and area swept columns.
#' @param ... Additional arguments passed to methods.
#'
#' @export
biomass_index <- function(x, ...) {
  UseMethod("biomass_index")
}

#' @rdname biomass_index
#' @param area_swept Numeric vector of area swept (e.g., km squared).
#' @param catch Optional numeric vector of catch. If provided with `effort`,
#'   CPUE is computed via [cpue()].
#' @param effort Optional numeric vector of effort. Required if `catch` is
#'   provided.
#' @export
biomass_index.numeric <- function(
  x,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  rlang::check_dots_used()

  if (is.null(x) && (!is.null(catch) && !is.null(effort))) {
    x <- cpue(catch, effort, ...)
  }

  if (is.null(x)) {
    stop("Must provide either 'x' (cpue) or both 'catch' and 'effort'.")
  }

  validate_numeric_inputs(x = x, area_swept = area_swept)

  x * area_swept
}

#' @rdname biomass_index
#' @param cpue_col Name of the CPUE column (default `"cpue"`).
#' @param area_col Name of the area swept column (default `"area_swept"`).
#' @export
biomass_index.data.frame <- function(
  x,
  cpue_col = "cpue",
  area_col = "area_swept",
  ...
) {
  rlang::check_dots_used()

  if (!cpue_col %in% names(x)) {
    stop("Column '", cpue_col, "' not found in data frame.", call. = FALSE)
  }
  if (!area_col %in% names(x)) {
    stop("Column '", area_col, "' not found in data frame.", call. = FALSE)
  }

  biomass_index(x[[cpue_col]], x[[area_col]])
}
```

Test it:

```{r}
document()
load_all()

# Vector method
biomass_index(c(10, 20), c(5, 8))

# Data frame method
survey_data <- data.frame(
  cpue = c(10, 20, 30),
  area_swept = c(5, 8, 6)
)

biomass_index(survey_data)
```

::: 
-->

## When to Use OOP in Packages

S3 classes are worth reaching for when:

- Your function returns complex results that benefit from custom `print`/`summary`/`plot` methods
- You want other packages to be able to extend your work with new methods
- You want to create a consistent interface when your users want to do the same
  things with different types of input (e.g., numeric vectors, data frames, model fits)

They are probably not needed when:

- A simple named list or data frame communicates the result clearly
- There is only one way to display or summarize the output
- The function returns a scalar or vector with obvious meaning

## Recap

- R has multiple OOP systems (S3, S4, R6) - S3 is the most common and covers most use cases
- S3 objects are regular R objects with a `class` attribute
- The same function (`print`, `summary`, `plot`) does different things for different classes -
  the object determines the behaviour, not the caller
- Constructors (`new_*`) build valid objects
- Write `print`, `summary`, and `plot` methods to make your classes pleasant to work with
- Create custom generics with `UseMethod()` for domain-specific operations
- Test classes with `expect_s3_class()` and snapshot tests for print output
