---
title: Testing with testthat - Part 2
format: html
---

### Updating snapshots

We have added better error messages with our input validations.
Because when we recorded the snapshots the error message was different, now when we run the tests, these snapshot tests will fail. 
We will need to update our snapshots.


```{r}
test()
```

We will get warning messages indicating that the snapshots have changed.

### Review snapshot changes

```{r}
# After modifying output, review changes
snapshot_review()
snapshot_accept()
```

## Writing "Clean" Tests

Tests should not leave behind any evidence that they were there. Use `withr` to keep things local.

`withr` has functions with the `local_` prefix that keeps changes local to the current environment. 
That environment can be a function, or it can be a test. 

There are also `with_` functions that are good for executing a small snippet of code with a modified state.

### Add withr to Suggests

```{r}
use_package("withr", type = "Suggests")
```

### Testing with package options

We updated `cpue()` to use a package option for the function verbosity. 
Remember our updated `R/cpue.R`:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)

  validate_numeric_inputs(catch = catch, effort = effort)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = catch / effort,
    log = log(catch / effort)
  )

  raw_cpue * gear_factor
}
```

Now test with `withr::local_options()` to temporarily change options:

In `tests/testthat/test-cpue.R`

```{r}
test_that("cpue uses verbosity when option set to TRUE", {
  withr::local_options(fishr.verbose = TRUE) # will be reset when this test_that block finishes

  expect_snapshot(cpue(100, 10))
})

test_that("cpue is not verbose when option set to FALSE", {
  withr::local_options(fishr.verbose = FALSE) # will be reset when this test_that block finishes

  expect_silent(cpue(100, 10))
})

test_that("cpue verbosity falls back to FALSE when not set", {
  withr::with_options(
    list(fishr.verbose = NULL), # will be reset as soon as this code block executes
    expect_no_message(cpue(100, 10))
  )
})
# Options automatically restored after each test
```

### Your turn

Add the same `verbose` argument to `biomass_index()` and add tests. Make sure
to update the documentation for `biomass_index()` with the new argument.

```{r}
#' Calculate Biomass Index
#'
#' Calculates biomass index from CPUE and area swept. Can optionally
#' compute CPUE from catch and effort data.
#'
#' @param cpue Numeric vector of CPUE values. If `catch` and `effort` are
#'   provided, this is computed automatically.
#' @param area_swept Numeric vector of area swept (e.g., km²)
#' @param catch Optional numeric vector of catch. If provided with `effort`,
#'   CPUE is computed via `cpue()`.
#' @param effort Optional numeric vector of effort. Required if `catch` is
#'   provided.
#' #' @param verbose Logical; print processing info? Default from
#'   `getOption("fishr.verbose", FALSE)`.
#' @param ... Additional arguments passed to `cpue()` when computing from
#'   catch and effort (e.g., `method`, `gear_factor`).
#'
#' @return A numeric vector of biomass index values
#' @export
#'
#' @examples
#' # From pre-computed CPUE
#' biomass_index(cpue = 10, area_swept = 5)
#'
#' # Compute CPUE on the fly
#' biomass_index(area_swept = 5, catch = 100, effort = 10)
#'
#' # Pass method through to cpue()
#' biomass_index(
#'   area_swept = 5,
#'   catch = c(100, 200),
#'   effort = c(10, 20),
#'   method = "log"
#' )
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  verbose = getOption("fishr.verbose", default = FALSE),
  ...
) {
  rlang::check_dots_used()

  if (is.null(cpue) && (!is.null(catch) && !is.null(effort))) {
    cpue <- cpue(catch, effort, verbose = verbose, ...)
  }

  if (is.null(cpue)) {
    stop("Must provide either 'cpue' or both 'catch' and 'effort'.")
  }

  validate_numeric_inputs(cpue = cpue, area_swept = area_swept)

  if (verbose) {
    message("calculating biomass index for ", length(area_swept), " records")
  }

  cpue * area_swept
}
```

```{r}
use_test("biomass")
```

In `tests/testthat/test-biomass.R`

```{r}
test_that("biomass_index uses verbosity when set as an option", {
  withr::local_options(fishr.verbose = TRUE)

  expect_snapshot(biomass_index(cpue = 5, area_swept = 100))
})

test_that("biomass_index verbosity falls back to FALSE when not set", {
  withr::local_options(fishr.verbose = NULL)

  expect_no_message(biomass_index(cpue = 5, area_swept = 100))
})
# Options automatically restored after each test
```


### Other withr helpers

`withr` provides many helpers for keeping tests isolated:

- `local_tempdir()` / `with_tempdir()` - create temporary directories
- `local_tempfile()` / `with_tempfile()` - create temporary files
- `local_envvar()` / `with_envvar()` - temporarily set environment variables
- `local_dir()` / `with_dir()` - temporarily change working directory
- `defer()` - register custom cleanup code

```{r}
test_that("example with temporary file", {
  temp_file <- withr::local_tempfile(lines = c("100,10", "200,20"))

  # Test something with temp_file
  lines <- readLines(temp_file)
  expect_length(lines, 2)
})
# temp_file automatically deleted after test
```

### Quiz

Can anyone think of somewhere we have altered the global state in our testing, 
and how we might address it?

::: {.callout-tip collapse="true"}

## Answer

In `tests/testthat/helper.R`: `set.seed()` sets the seed globally. We want to 
create the data frame the same every time, so need the same random seed, but
want to keep that change local.

```{r}
# tests/testthat/helper.R

# Helper function to generate sample fishing data
generate_fishing_data <- function(n = 10) {
  withr::local_seed(67)
  data.frame(
    catch = runif(n, 10, 500),
    effort = runif(n, 1, 20),
    gear_factor = runif(n, 1, 5)
  )
}
```

:::


## Test Coverage with covr

Once you have a test suite, a natural question is: how much of my code is actually being tested? 
The `covr` package answers this by tracking which lines of your package code are executed when your tests run.

### Set up covr

Add covr to `Suggests` in your DESCRIPTION

```{r}
use_package("covr", type = "Suggests")
```

### Checking coverage

`package_coverage()` runs your tests and records which lines were hit:

```{r}
library(covr)

cov <- package_coverage()
cov
```

This prints a per-file coverage percentage. 
To get a single overall number:

```{r}
percent_coverage(cov)
```

### Finding untested code

The most actionable output from covr is the list of lines with zero coverage:

```{r}
zero_coverage(cov)
```

This returns a data frame showing exactly which lines have no test exercising them. 
In RStudio, the results appear as markers you can click to jump to the uncovered code.

### Interactive report

For a visual overview, `report()` opens an HTML page where you can browse file-by-file coverage with highlighted source:

```{r}
report(cov)
```

### Interactive checking

```{r}
devtools::test_coverage_active_file()
```

Nice to set a keyboard shortcut for this: `Ctrl/Cmd + R`

### Excluding lines from coverage

Some lines are not worth testing - for example, a `print` method or an interactive-only code path. 
You can exclude them with a special comment:

```{r}
if (interactive()) {
  # nocov start
  browse_results(x)
} # nocov end
```

Or exclude a single line:

```{r}
stop("Column must be numeric") # nocov
```

### Coverage in CI (optional content)

To also run coverage automatically in CI, add the GitHub Actions workflow:

```{r}
usethis::use_github_action("test-coverage")
```

By default the workflow uploads results to [Codecov](https://about.codecov.io/), which requires a (free) Codecov account linked to your GitHub repo. 
Codecov provides a dashboard that tracks coverage over time and can comment on pull requests with coverage diffs.

If you use Codecov, you will need to add a `CODECOV_TOKEN` as a repository secret in GitHub. 
To get your token, sign in to [codecov.io](https://about.codecov.io/) with your GitHub account, navigate to your repository, and copy the upload token. 
Then in your GitHub repo, go to **Settings > Secrets and variables > Actions** and add a new repository secret named `CODECOV_TOKEN` with that value. 
The workflow generated by `use_github_action("test-coverage")` already references this secret, so once it is set uploads will work automatically.

If you don't want to set up a Codecov account, you can edit the generated workflow file (`.github/workflows/test-coverage.yaml`) to just print coverage to the Actions log without uploading. 
The default workflow has a "Test coverage" step that already calls `print(cov)`, followed by a `codecov/codecov-action` step that uploads. 
Remove the upload step and the `covr::to_cobertura()` call so the "Test coverage" step looks like this:

```yaml
      - name: Test coverage
        run: |
          cov <- covr::package_coverage(
            quiet = FALSE,
            clean = FALSE,
            install_path = file.path(
              normalizePath(Sys.getenv("RUNNER_TEMP"), winslash = "/"),
              "package"
            )
          )
          print(cov)
        shell: Rscript {0}
```

Coverage results will appear directly in the Actions log each time the workflow runs. 
You still get visibility into coverage on every push and PR - you just don't get the Codecov dashboard or PR comments.

### A note on coverage targets

High coverage is useful as a guide for finding gaps, but chasing 100% coverage is rarely worthwhile. 
Coverage tells you which lines were *executed*, not whether the tests actually verify correct behavior. 
A test that runs code without meaningful assertions inflates coverage without adding value. 
Focus on covering the important logic paths and edge cases rather than hitting a number.
