---
title: "Data in Packages"
subtitle: "Advanced R Package Development"
format:
  revealjs:
    height: 900
    width: 1600
    theme: [moon, slides.scss]
    highlight-style: a11y
    slide-number: true
    chalkboard: true
    footer: "[{{< fa house >}}](../index.html)"
execute:
  eval: false
  echo: true
---

## Outline

1. Why include data in packages?
2. Static exported data
3. Fetching live data with httr2
4. Internal data

# Why Include Data in Packages?

## Data is already everywhere in R

`iris`, `mtcars`, `airquality` - you have been using package data all along.

Beyond bundling examples, packages can:

- Ship **reference tables** that functions need internally
- Bundle **example datasets** that make documentation self-contained
- Act as **data packages** whose sole purpose is distributing data
- Provide **API functions** that fetch live data on demand

::: {.fragment}
Today we cover the first, second, and fourth of these in `fishr`.
:::

## Types of package data

| Location | Created by | Exported? | Use when |
|---|---|---|---|
| `data/*.rda` | `use_data()` | Yes | Example data, datasets users work with directly |
| `R/sysdata.rda` | `use_data(..., internal = TRUE)` | No | Lookup tables, constants used by functions |
| `inst/extdata/` | Manual copy | No | Non-R formats (CSV, GeoJSON, shapefiles) |
| API function | `httr2` | Yes | Live data that changes over time |

# Static Exported Data

## The workflow

Keep reproducible scripts in `data-raw/` that generate your data objects. This script is the ground truth for how your data was created.

```{r}
usethis::use_data_raw("horsefly_sample")
```

::: {.fragment}
Creates `data-raw/horsefly_sample.R` and adds `data-raw/` to `.Rbuildignore` - the directory is excluded from the installed package, but commit the scripts inside it.
:::

## Build and save the dataset

```{r}
# data-raw/horsefly_sample.R

horsefly_sample <- data.frame(
  set_number = c(1L, 2L, 3L, 4L, 5L),
  catch      = c(3L, 0L, 5L, 2L, 4L),
  effort     = c(1L, 1L, 1L, 1L, 1L),
  stringsAsFactors = FALSE
)

usethis::use_data(horsefly_sample, overwrite = TRUE)
```

::: {.fragment}
Run the script once to create `data/horsefly_sample.rda`:

```{r}
source("data-raw/horsefly_sample.R")
```
:::

## Document exported data

Exported datasets need documentation just like functions. Create `R/data.R`:

```{r}
use_r("data")
```

```r
#' Rainbow Trout gill net survey data from Horsefly Lake
#'
#' Catch and effort from a fictional gill net survey of Rainbow Trout
#' (*Oncorhynchus mykiss*) at Horsefly Lake, British Columbia.
#'
#' @format A data frame with columns:
#' - `set_number` Integer. Net set identifier.
#' - `catch` Integer. Number of fish caught per set.
#' - `effort` Integer. Number of net-nights per set.
#' @source Fictional data for illustration purposes.
#' @examples
#' cpue(horsefly_sample)
"horsefly_sample"
```

## Try it out

```{r}
document()
load_all()
```

::: {.fragment}
The dataset appears in `data(package = "fishr")`.

```{r}
horsefly_sample

?horsefly_sample

cpue(horsefly_sample)
```

Because `horsefly_sample` already has `catch` and `effort` columns, it passes directly to `cpue()`.
:::

# Fetching Live Data with httr2

## Package data vs package functionality

`fetch_kluane_data()` is not a dataset - it is a function that returns data.

- Lives in `R/`, documented with roxygen2, tested like any other function
- The Kluane Lake Trout data is updated regularly, so a function that fetches the current snapshot is more useful than a static copy bundled at build time

## Add httr2 as a dependency

```{r}
usethis::use_package("httr2")
```

Adds `httr2` to `Imports` in `DESCRIPTION`.

## The Government of Canada Open Data API

The Kluane Lake Trout dataset is published via CKAN:

- **Dataset page:** [open.canada.ca](https://open.canada.ca/data/en/dataset/307fce02-71cf-4902-b485-57a70cfafe2b)
- **API endpoint:** `https://open.canada.ca/data/en/api/3/action/datastore_search`
- **Resource ID:** `af1e5730-34bd-4314-831c-1d940d99e1a7`

The response is JSON with a `result.records` array where each element is one row.

## Building fetch_kluane_data()

```{r}
fetch_kluane_data <- function(limit = 2000) {
  req <- httr2::request(
    "https://open.canada.ca/data/en/api/3/action/datastore_search"
  )
  req <- httr2::req_url_query(
    req,
    resource_id = "af1e5730-34bd-4314-831c-1d940d99e1a7",
    limit = limit
  )
  resp <- httr2::req_perform(req)
  httr2::resp_check_status(resp)

  records <- httr2::resp_body_json(resp, simplifyVector = TRUE)$result$records
  records <- records[-1, ] # first row is a bilingual EN/FR header, not data

  data.frame(
    lake           = records$Lake,
    year           = as.integer(records$Year),
    set_number     = as.integer(records[["Set Number"]]),
    species        = records$Species,
    fork_length_mm = as.numeric(records[["Fork Length (millimetres)"]]),
    weight_g       = as.numeric(records[["Weight (grams)"]]),
    stringsAsFactors = FALSE
  )
}
```

## httr2 step by step

- `request()` - create a request object pointing at the base URL
- `req_url_query()` - append query parameters (`?resource_id=...&limit=...`)
- `req_perform()` - execute the HTTP GET and return a response object
- `resp_check_status()` - raise an informative error on 4xx/5xx responses
- `resp_body_json(..., simplifyVector = TRUE)` - parse JSON and coerce the array of records directly to a data frame

## Try it out

```{r}
load_all()

kluane <- fetch_kluane_data()
head(kluane)
nrow(kluane)
```

## Connecting to cpue()

```{r}
fish <- kluane[kluane$species != "no fish", ]

survey_df <- data.frame(
  catch  = as.integer(table(fish$set_number)),
  effort = 1L
)

lt_cpue <- cpue(survey_df)
summary(lt_cpue)
```

::: {.fragment}
`cpue()` is a generic with a `data.frame` method - passing a data frame with `catch` and `effort` columns dispatches automatically. Real-world data flows through the same function without any changes to it.
:::

## Testing functions that call HTTP APIs

Testing functions that make real HTTP requests requires care - tests should not depend on an external server being available.

The [vcr](https://docs.ropensci.org/vcr/) and [httptest2](https://enpiar.com/httptest2/) packages solve this by recording real HTTP interactions and replaying them during tests, keeping your test suite fast and self-contained.

# Internal Data

## What is internal data for?

Some data belongs to the package but should not be visible to users.

Good candidates: lookup tables, constants, calibration coefficients.

::: {.fragment}
Internal data powers package behaviour without appearing in the user's namespace.

It lives in `R/sysdata.rda`, not `data/`.
:::

## Creating internal data

```{r}
use_data_raw("gear-types")
```

```{r}
# data-raw/gear_types.R
gear_types <- data.frame(
  gear_type   = c("nordic_gillnet", "sinking_longline", "fyke_net", "electrofishing"),
  gear_factor = c(1.0, 0.72, 1.35, 0.45),
  description = c(
    "Nordic multi-mesh gillnet (standard reference gear)",
    "Sinking baited longline",
    "Passive fyke net trap",
    "Electrofishing unit"
  ),
  stringsAsFactors = FALSE
)

usethis::use_data(gear_types, internal = TRUE, overwrite = TRUE)
```

::: {.fragment}
```{r}
source("data-raw/gear_types.R")
```

Every function in the package can access `gear_types` directly, but it does not appear in `ls()` for users.
:::

## Using internal data in a function

Replace the numeric `gear_factor` argument with a named `gear_type`:

```{r}
cpue.numeric <- function(
  x,
  effort,
  gear_type = "nordic_gillnet",
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE),
  ...
) {
  if (!gear_type %in% gear_types$gear_type) {
    stop(
      "`gear_type` must be one of: ",
      paste(gear_types$gear_type, collapse = ", "),
      call. = FALSE
    )
  }

  gear_factor <- gear_types$gear_factor[gear_types$gear_type == gear_type]
  # ...
}
```

## Dynamic error messages

The error message is generated from the table at runtime:

```{r}
load_all()

cpue(100, 10)                          # nordic_gillnet -> 10
cpue(100, 10, gear_type = "fyke_net")  # factor 1.35 -> 13.5
cpue(100, 10, gear_type = "trapnet")   # error: lists all valid types
```

::: {.fragment}
Add a gear type to the table, re-run the script, and it appears in the error automatically - no manual string editing required.
:::

# Your Turn

## Exercise

**Goal:** extend the gear type table, verify the error updates automatically, fix the tests.

1. Open `data-raw/gear_types.R` and add a row for `"trap_net"` with a gear factor of `1.1`

2. Re-run the script and reload:

```{r}
source("data-raw/gear_types.R")
load_all()
```

3. Confirm the new gear type works and the error now lists `"trap_net"`:

```{r}
cpue(100, 10, gear_type = "trap_net") # should return 110
cpue(100, 10, gear_type = "unknown")  # error should list trap_net
```

4. Run `check()` - changing `gear_factor` to `gear_type` broke existing tests. Fix them.
