---
title: "Advanced Documentation"
subtitle: "Day 2, Session 3 - Advanced R Package Development"
format:
  revealjs:
    height: 900
    width: 1600
    theme: [moon, slides.scss]
    highlight-style: a11y
    slide-number: true
    chalkboard: true
    footer: "[{{< fa house >}}](../index.html)"
execute:
  eval: false
  echo: true
---

## Outline

1. The duplication problem
2. `@inheritParams`
3. `@inheritDotParams`
4. Markdown formatting
5. Cross-references

# The Duplication Problem

## Two functions, overlapping parameters

`cpue()` and `biomass_index()` share `catch`, `effort`, and `verbose`.

::: columns
::: {.column width="50%"}
**`cpue()`**

```{r}
#| code-line-numbers: "|1-4,9-10"
#' @param catch Numeric vector of
#'   catch (e.g., kg).
#' @param effort Numeric vector of
#'   effort (e.g., hours).
#' @param gear_factor Numeric scalar for gear 
#'   standardization (default 1)
#' @param method Character; 
#'   one of `"ratio"` (default) or `"log"`.
#' @param verbose Logical; Default from
#'   `getOption("fishr.verbose", FALSE)`.
```
:::

::: {.column width="50%"}
**`biomass_index()`**

```{r}
#| code-line-numbers: "|4-9|10-11|4-11"
#' @param cpue Numeric vector of CPUE values. 
#' @param area_swept Numeric vector of area 
#'   swept (e.g., km^2^)
#' @param catch Numeric vector of
#'   catch (e.g., kg).
#' @param effort Numeric vector of
#'   effort (e.g., hours).
#' @param verbose Logical; Default from
#'   `getOption("fishr.verbose", FALSE)`.
#' @param ... Additional arguments 
#'   passed to `cpue()`.
```
:::
:::

::: {.fragment}
**Two problems:**

1. If the descriptions ever change, you have two places to update
2. `@param ...` is vague - users cannot tell which arguments are accepted and passed to `cpue()`
:::

# `@inheritParams`

## Copy parameter docs from another function

`@inheritParams` pulls `@param` entries from a source function into the current one.

```{r}
#| code-line-numbers: "|6"
#' Calculate Biomass Index
#'
#' @param cpue Numeric vector of CPUE values. If NULL, computed from `catch`
#'   and `effort`.
#' @param area_swept Numeric vector of area swept (e.g., km^2^).
#' @inheritParams cpue
#' @param ... Additional arguments passed to `cpue()`.
#' @export
```

::: {.fragment}

Render documentation and open the help page:

```{r}
document()
?biomass_index
```
:::

## What gets inherited?

roxygen2 copies `@param` entries from `cpue()` where the parameter:

::: {.incremental}
- appears in `biomass_index()`'s signature, **and**
- is not already documented in `biomass_index()`
:::

::: {.fragment}
- `catch`, `effort`, and `verbose` are inherited. 
- `gear_factor` and `method` are skipped - they are not explicit parameters of `biomass_index()`.
:::

::: {.fragment}
`@inheritParams` also works across packages: `@inheritParams pkg::function`
:::

::: {.notes}
Point out that @inheritParams can chain: if C inherits from B which inherits from A, roxygen2 resolves the full chain. The descriptions stay in sync automatically - change the docs in cpue() and biomass_index() picks them up on the next document() call.
:::

# `@inheritDotParams`

## Make `...` informative

`@param ...` tells users nothing. 

```{r}
#' @param ... Additional arguments passed to `cpue()`.
```

::: {.fragment}
What are the additional arguments? What do they do? \
The user has to  go to the `?cpue()` help and figure it out for themselves.
:::

## `@inheritDotParams` {auto-animate="true"}

generates a formatted description that names the forwarded arguments and links their documentation.

```{r}
#| code-line-numbers: "7"
#' Calculate Biomass Index
#'
#' @param cpue Numeric vector of CPUE values. If NULL, computed from `catch`
#'   and `effort`.
#' @param area_swept Numeric vector of area swept (e.g., km^2^).
#' @inheritParams cpue
#' ... Additional arguments passed to `cpue()`.
```

## `@inheritDotParams` {auto-animate="true"}

generates a formatted description that names the forwarded arguments and links their documentation.

```{r}
#| code-line-numbers: "7"
#' Calculate Biomass Index
#'
#' @param cpue Numeric vector of CPUE values. If NULL, computed from `catch`
#'   and `effort`.
#' @param area_swept Numeric vector of area swept (e.g., km^2^).
#' @inheritParams cpue
#' @inheritDotParams cpue -catch -effort
```

::: {.fragment}
The `-catch -effort` syntax **excludes** those parameters - they are already documented explicitly above.

What remains (`gear_factor`, `method`) is rendered as a formatted list linking back to `?cpue`.
:::

## `@inheritDotParams`

The `...` entry in the help page now reads something like:

\

**`...`** Arguments passed to `cpue()`:\
&emsp;`gear_factor` Numeric adjustment for gear standardisation...\
&emsp;`method` One of "ratio" or "log"...


::: {.notes}
Before @inheritDotParams, a user had to navigate to ?cpue, figure out which arguments apply to biomass_index(), and mentally filter. Now they get the answer on the same help page.
:::

# Markdown Formatting

## Inline formatting

Standard Markdown works in all roxygen2 text fields: `@param`, `@return`, `@description`, `@details`.

```{r}
#' @param area_swept Numeric vector of area swept in **km^2^**.
#' @return A numeric vector the same length as `cpue`. Returns `NA` for any
#'   row where either input is `NA`.
```

::: {.fragment}
- Backticks (\`\`) render as inline code: `cpue`, `NA`, `TRUE`
- `**text**` renders as **bold**
- `*text*` renders as *italic*
- `^text^` renders as superscript: km^2^
:::

## Lists

```{r}
#' @details
#' Two modes of use:
#'
#' - Provide `cpue` directly for a simple calculation.
#' - Provide `catch` and `effort` to compute CPUE first, then scale by area.
```

::: {.fragment}
::: {.notebox}
The blank line between the tag and the list is required. Without it, roxygen2 treats the hyphens as literal text.
:::
:::

# Cross-References

## Linking to functions

The most common form: function name followed by `()` in square brackets `[]`.

```{r}
#| code-line-numbers: "2"
#' Calculates biomass index from CPUE and area swept.
#' See [cpue()] for computing CPUE values from raw catch and effort.
```

::: {.fragment}
The `()` suffix tells roxygen2 this is a function link. It generates a code-formatted hyperlink to `?cpue`.
:::

## Linking across packages

Use the `::` operator inside the brackets:

```{r}
#' @param data A data frame. See [dplyr::filter()] for subsetting rows.
```

::: {.fragment}
This creates a link that opens `?dplyr::filter`.

R CMD check may produce a NOTE: "Undeclared package 'dplyr' in Rd xrefs". Adding the package to `Suggests` in your DESCRIPTION silences it - you do not need `Imports`.
:::

## Custom link text and `@seealso`

To display different text than the function name:

```{r}
#' Returns the same structure as [the CPUE function][cpue()].
```

::: {.fragment}
Use this sparingly - the function name itself is usually the most informative link text.
:::

::: {.fragment}
`@seealso` renders in its own section at the bottom of the help page:

```{r}
#' @seealso [cpue()] to compute CPUE values from raw catch and effort.
```

Use it for functions that are related but not directly referenced in the body of the docs.
:::

## Summary

| Syntax | Result |
|--------|--------|
| `[cpue()]` | Link to `?cpue` in this package |
| `[pkg::cpue()]` | Link to `?pkg::cpue` in another package |
| `[some text][cpue()]` | "some text" linking to `?cpue` |
| `@inheritParams cpue` | Copy matching `@param` entries from `cpue()` |
| `@inheritDotParams cpue -x` | Document `...` as forwarded args, excluding `x` |

::: {.fragment}
Use links freely in `@description`, `@details`, `@param`, `@return`, and `@seealso`. A well-linked help page lets users navigate the package without leaving the help system.
:::
