---
title: "Function Design Best Practices"
subtitle: "Advanced R Package Development"
format:
  revealjs:
    height: 900
    width: 1600
    theme: [moon, slides.scss]
    highlight-style: a11y
    slide-number: true
    chalkboard: true
    footer: "[{{< fa house >}}](../index.html)"
execute:
  eval: false
  echo: true
---

## Outline

1. Why function design matters
2. Naming conventions
3. Pure functions vs side effects
4. Default arguments and `match.arg`
5. The ellipsis (`...`)
6. Managing dependencies
7. DRY: Don't Repeat Yourself
8. Function composition

# Why Function Design Matters

## What makes good code?

- Readable: easy to understand what it does
- Correct: does what you intend
- Maintainable: easy to change

::: {.fragment}
From the [Tidyverse design guide](https://design.tidyverse.org/):

- Programming is a task performed by humans
- Reduce cognitive load with consistent design
- Make your functions and systems composable
- Think about others who are not like us
:::

::: {.notes}
Today we refactor our fishr functions with these principles in mind. The goal is not perfection - it's making functions easier to use correctly and harder to use incorrectly.
:::

# Naming Conventions

## General guidelines

- Use verbs to ascribe an action
- Use consistent style (`snake_case`)
- Consider short prefixes to unify package functions
- Don't be afraid to be verbose
- Avoid conflict with existing functions

## snake_case for everything

```{r}
# Good
calculate_area <- function(length, width) {}
lib_summary <- function() {}

# Avoid
calculateArea <- function(Length, Width) {}
```

## Functions as verbs

```{r}
# Good - functions do things
calculate_cpue()
filter_by_year()
plot_abundance()

# Avoid
result() # What result?
fish_data() # Is this a function or an object?
```

::: {.fragment}
Note: Sometimes you have exceptions. `cpue` us a well-established domain abbreviation. The next slide covers this.
:::

## Objects as nouns

```{r}
# Good - objects are things
survey_data <- read_csv("survey.csv")
model_fit <- lm(y ~ x, data = df)

# Avoid
doReadingCSV <- read_csv("survey.csv")
```

## Code style

Consistent style makes code easier to read.

```{r}
# Inconsistent
x = 1 + 2
y = sum(c(1, 2, 3))

# Consistent
x <- 1 + 2
y <- sum(c(1, 2, 3))
```

::: {.fragment}
The `air` formatter (hopefully already configured in this project) handles this automatically on save.
:::

## Naming in fishr

Our functions follow these conventions:

- `cpue()` - short, but a domain-standard abbreviation well understood by fisheries scientists
- `biomass_index()` - descriptive, easy to predict

::: {.fragment}
Domain-standard abbreviations are an example of an exception to be verbose.
:::

# Pure Functions vs Side Effects

## Two kinds of functions

**Pure function:**

1. Always returns the same output for the same inputs
2. Has no side effects

```{r}
# Pure
add <- function(x, y) {
  x + y
}

# Not pure - depends on global state
multiplier <- 2
scale <- function(x) {
  x * multiplier
}
```

## Functions with side effects

```{r}
# Side effect: modifies global state
counter <- 0
increment <- function() {
  counter <<- counter + 1
}

# Side effects: reads from disk, prints output
load_data <- function(path) {
  cat("Loading", path, "\n")
  read.csv(path)
}
```

::: {.fragment}
Side effects are sometimes necessary, but prefer pure functions when possible. Easier to test and reason about.
:::

## cpue annotated

```{r}
cpue <- function(catch, effort, gear_factor = 1, verbose = FALSE) {
  # Side effect: prints a message to the console
  if (verbose) {
    message("Processing ", length(catch), " records")
  }

  # Pure calculation
  raw_cpue <- catch / effort
  raw_cpue * gear_factor
}
```

The calculation itself is pure. The `message()` call is a side effect.

## Package-wide options

R has a global key-value store that packages can use for user-configurable defaults:

```{r}
# Set an option
options(fishr.verbose = TRUE)

# Get an option (with a fallback default)
getOption("fishr.verbose", default = FALSE)
```

::: {.fragment}
By convention, package options use the package name as a prefix to avoid conflicts: `fishr.verbose`, not just `verbose`.
:::

## getOption() in the function signature

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  verbose = getOption("fishr.verbose", default = FALSE)
) {
  if (verbose) {
    message("Processing ", length(catch), " records")
  }

  raw_cpue <- catch / effort
  raw_cpue * gear_factor
}
```

::: {.fragment}
Users can set `options(fishr.verbose = TRUE)` once and all fishr functions pick it up. They can still override per call: `cpue(100, 10, verbose = TRUE)`.
:::

## Side effects to avoid

```{r}
# Don't do this - writes a file without asking
bad_cpue <- function(catch, effort) {
  result <- catch / effort
  write.csv(data.frame(cpue = result), "cpue_log.csv")
  result
}

# Don't do this - changes how all numbers display globally
bad_summary <- function(x) {
  options(digits = 2)
  summary(x)
}
```

::: {.fragment}
Return values and let the caller decide what to do with them.
:::

# Default Arguments and match.arg

## Three types of arguments

```{r}
my_function <- function(
  required_arg, # Required: no default
  optional_arg = TRUE, # Optional: has a default
  ... # Dots: passed to other functions
) {}
```

## Adding a method argument to cpue

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(
    method,
    ratio = catch / effort,
    log = log(catch / effort)
  )

  raw_cpue * gear_factor
}
```

## How match.arg works

`match.arg()` takes the first element of the default vector when the user doesn't supply a value, and gives a clear error for invalid input:

```{r}
load_all()

cpue(100, 10) # Uses default ("ratio")
cpue(100, 10, method = "log") # Explicit
cpue(100, 10, method = "median") # Error: invalid value
```

::: {.fragment}
The vector in the default `c("ratio", "log")` is both the documentation and the validation list.
:::

## Document and check

```{r}
document()
check()
```

::: {.fragment}
Adding `method` to the verbose message changes its format - any snapshot test for the verbose output will fail with a mismatch. Accept the updated snapshot:

```{r}
testthat::snapshot_accept()
```

:::

# The Ellipsis (...)

## Passing arguments through with ...

The ellipsis lets a function accept extra arguments and pass them to another function:

```{r}
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  if (is.null(cpue) && (!is.null(catch) && !is.null(effort))) {
    cpue <- cpue(catch, effort, ...)
  }

  if (is.null(cpue)) {
    stop("Must provide either 'cpue' or both 'catch' and 'effort'.")
  }

  cpue * area_swept
}
```

## Demo

```{r}
load_all()

# Pre-computed CPUE
biomass_index(cpue = 10, area_swept = 5)

# Compute on the fly
biomass_index(area_swept = 5, catch = 100, effort = 10)

# Pass method through to cpue() via ...
biomass_index(
  area_swept = 5,
  catch = c(100, 200),
  effort = c(10, 20),
  method = "log"
)
```

::: {.notes}
Note: we write area_swept = 5 even though it's the second argument. Because we're skipping cpue, positional matching would put 5 into cpue instead. When you skip optional arguments, name them.
:::

## Why cpue = NULL?

`cpue = NULL` signals "not provided" clearly:

- `is.null()` is a reliable check for "argument not given"
- If we used `cpue = 0`, there's no way to tell whether the user passed zero intentionally or just didn't provide a value

## Pitfalls of ...

- Arguments passed through `...` must be named - positional matching doesn't work through the ellipsis
- Misspelled arguments are silently absorbed: `cpue(100, 10, mthod = "log")` won't error
- Errors from inner functions can be hard to trace

::: {.fragment}
Solution: `rlang::check_dots_used()` catches unused dots before they cause confusion.
:::

# Managing Dependencies

## Declaring a dependency

```{r}
use_package("rlang")
```

::: {.fragment}
Adds `rlang` to `Imports` in `DESCRIPTION`. Call functions using `::`:

```{r}
format_catch <- function(catch) {
  if (!is.numeric(catch)) {
    rlang::abort("`catch` must be a numeric vector.")
  }
  round(catch, 2)
}
```

`::` is explicit, avoids namespace conflicts, and makes clear which package each function comes from.
:::

## Importing frequently-used functions

For functions called throughout a package, `::` on every call is verbose:

```{r}
use_import_from("rlang", "abort")
```

::: {.fragment}
Inserts `@importFrom rlang abort` into your package roxygen block and updates `NAMESPACE` after `document()`.

`abort()` is then available across your package without the `rlang::` prefix.
:::

## Catching unused dots

```{r}
biomass_index <- function(
  cpue = NULL,
  area_swept,
  catch = NULL,
  effort = NULL,
  ...
) {
  rlang::check_dots_used()

  if (is.null(cpue) && (!is.null(catch) && !is.null(effort))) {
    cpue <- cpue(catch, effort, ...)
  }

  if (is.null(cpue)) {
    stop("Must provide either 'cpue' or both 'catch' and 'effort'.")
  }

  cpue * area_swept
}
```

# DRY: Don't Repeat Yourself

## The problem with repetition

```{r}
# Repetitive
sum_of_squares_2 <- 1^2 + 2^2
sum_of_squares_3 <- 1^2 + 2^2 + 3^2
sum_of_squares_4 <- 1^2 + 2^2 + 3^2 + 4^2
```

::: {.fragment}
Problems:

- Change the formula and you must update every line
- Easy to introduce inconsistencies
- Hard to read and maintain
:::

## Refactor to use functions

```{r}
# DRY version
square <- function(x) x^2

sum_of_squares <- function(n) {
  sum(square(seq_len(n)))
}

sum_of_squares(2)
sum_of_squares(3)
sum_of_squares(4)
```

Small, focused functions that do one thing well. Easier to test, read, and reuse.

## Identifying repetition

Look for:

- Copy-pasted code with small variations
- The same expression computed multiple times
- Long functions that do many separate things

::: {.fragment}
If a function exceeds ~20-30 lines, consider whether it is doing too many things.
:::

## Why validate inputs?

```{r}
cpue("one hundred", 10)
#> Error in catch/effort : non-numeric argument to binary operator
```

::: {.fragment}
Unhelpful - doesn't say which argument is the problem or what was expected.

Instead of copy-pasting the same check into every function, create a shared helper.
:::

## A validation helper

```{r}
# R/utils.R

#' @noRd
validate_numeric_inputs <- function(...) {
  args <- list(...)
  arg_names <- names(args)

  for (i in seq_along(args)) {
    if (!is.numeric(args[[i]])) {
      stop(
        "'",
        arg_names[i],
        "' must be numeric, got ",
        class(args[[i]])[1],
        ".",
        call. = FALSE
      )
    }
  }

  invisible(NULL)
}
```

`@noRd` tells roxygen2 not to generate a `.Rd` file - internal helpers don't need public docs.

## Add validation to cpue

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) {
  method <- match.arg(method)
  validate_numeric_inputs(catch = catch, effort = effort)

  if (verbose) {
    message("Processing ", length(catch), " records using ", method, " method")
  }

  raw_cpue <- switch(method, ratio = catch / effort, log = log(catch / effort))
  raw_cpue * gear_factor
}
```

## Verify the helper works

```{r}
load_all()

cpue(100, 10) # Good input
cpue("high", 10) # 'catch' must be numeric, got character.
biomass_index(cpue = "ten", area_swept = 5) # 'cpue' must be numeric...
```

## Where to put helpers

- **Same file as the exported function** - when the helper is specific to one function. Keeps related logic together.
- **`R/utils.R`** - when the helper is used across multiple functions in the package.

::: {.fragment}
Good rule of thumb: start with the helper in the same file. Move it to `utils.R` only when a second function needs it.
:::

# Function Composition

## Composed workflow

```{r}
load_all()

# Step by step
my_cpue <- cpue(catch = c(100, 200, 300), effort = c(10, 20, 30))
biomass_index(cpue = my_cpue, area_swept = 50)

# Or in one call, thanks to ... pass-through
biomass_index(area_swept = 50, catch = c(100, 200, 300), effort = c(10, 20, 30))

# With options passed through
biomass_index(
  area_swept = 50,
  catch = c(100, 200, 300),
  effort = c(10, 20, 30),
  method = "log"
)
```

## Composability is the payoff

- **Naming** - clear names make composed code readable
- **Pure calculations** - predictable functions are safe to chain
- **`match.arg`** - consistent interfaces reduce mistakes
- **`...`** - pass-through enables flexible composition
- **Helpers** - shared validation keeps behaviour consistent

# Your Turn

## Exercise

Add `verbose` support to `biomass_index()` using the same options pattern as `cpue()`:

1. Add `verbose = getOption("fishr.verbose", FALSE)` as an argument
2. When `verbose = TRUE`, print a message reporting how many records are being processed
3. Run `document()` and `check()`
4. Update any snapshot tests that now fail

::: {.fragment}
**Bonus:** write a test that confirms the message appears when `verbose = TRUE` and is silent by default.
:::
