---
title: "Git/GitHub for Collaborative Development"
subtitle: "Advanced R Package Development"
format:
  revealjs:
    height: 900
    width: 1600
    theme: [moon, slides.scss]
    highlight-style: a11y
    slide-number: true
    chalkboard: true
    footer: "[{{< fa house >}}](../index.html)"
execute:
  eval: false
  echo: true
---

## Outline

1. Git workflow review
2. Pull request workflows with `usethis`
3. Continuous integration with GitHub Actions
4. Code formatting with Air
5. Adding a function via PR

# Git Workflow Review

## Where we left off

Make sure all changes from the fundamentals session are committed and pushed:

```{bash}
git push origin main
```

::: {.notes}
Quick check - everyone should have a fishr package with `cpue()` and basic documentation, all pushed to GitHub. 
If anyone is behind, now is the time to catch up before we add CI.
:::

## The core loop

Every day of package development follows the same loop:

::: {.incremental}
1. Pull latest changes
2. Make a small, focused change
3. Test that it works
4. Commit with a clear message
5. Push
:::

::: {.fragment}
Pull requests add a **review step** between commit and merge to `main`.
:::

# Pull Request Workflows

## Collaborating

- Git is a *distributed* version control system
- Each collaborator has a full copy of the repository, including its history
- All collaborators can push changes to the remote repository (given they have permission)
- Changes made by collaborators can be pulled into your local repository

## Why pull requests?

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: pull-request
%%| fig-width: 14
%%| file: pull-request.mmd
```

Even when working alone, PRs are valuable:

::: {.incremental}
- Check code runs before it reaches `main`
- You get a diff to review your own/others' work
- A clean history of *why* changes were made
- Allows fearless experimentation
:::

::: {.fragment}
With collaborators, PRs also enable **code review** - the single biggest driver of code quality.
:::

## The `usethis` PR helpers

| Function | What it does |
|---|---|
| `pr_init("name")` | Create and switch to a new branch |
| `pr_push()` | Push branch, open PR on GitHub |
| `pr_pause()` | Stash work, return to `main` |
| `pr_resume("name")` | Return to a branch |
| `pr_fetch(123)` | Check out someone else's PR |
| `pr_finish()` | Clean up after merge |

## The PR workflow: overview

```{mermaid}
%%| eval: true
%%| echo: false
%%| label: pr-helpers
%%| fig-width: 16
%%| fig-height: 8

---
config:
  theme: base
  themeVariables:
    fontSize: 30px
---

flowchart LR
    A([main]) --> PI["pr_init()"] --> B([Branch])
    B --> C[Write & test]
    C --> PP["pr_push()"] --> D([GitHub PR])
    D --> E{Review}
    E --> C
    E -->|Merged| F([main updated])
    F --> PF["pr_finish()"] --> G([Clean state])

    style A fill:#2d6a4f,color:#fff,stroke:none
    style F fill:#2d6a4f,color:#fff,stroke:none
    style G fill:#2d6a4f,color:#fff,stroke:none
    style B fill:#1d3557,color:#fff,stroke:none
    style D fill:#457b9d,color:#fff,stroke:none
    style E fill:#e63946,color:#fff,stroke:none
    style C fill:#333,color:#fff,stroke:#666
    style PI fill:#f4a261,color:#000,stroke:none
    style PP fill:#f4a261,color:#000,stroke:none
    style PF fill:#f4a261,color:#000,stroke:none
```

## Starting a branch

```{r}
usethis::pr_init("my-branch-name")
```

::: {.fragment}
- Creates and checks out the branch immediately
- You are now isolated from `main`
- Branch names: short, descriptive, hyphenated
:::

::: {.notes}
A reader should be able to guess what the PR does from the branch name alone.
:::

## Pushing and opening a PR

```{r}
usethis::pr_push()
```

::: {.fragment}
- Pushes your branch to GitHub
- Opens your browser to the "Create pull request" page
- Pre-fills the branch comparison
:::

## Writing a good PR description

**Title:** Short, imperative - "Add biomass_index function"

**Body:**

- What problem does this solve?
- Any design decisions worth explaining?
- Anything reviewers should pay attention to?

::: {.fragment}
A good PR description is a gift to your future self and your collaborators.
:::

## After review and merge

Once the PR is approved and merged on GitHub:

```{r}
usethis::pr_finish()
```

::: {.fragment}
- Switches back to `main`
- Pulls the latest changes (including your merged work)
- Deletes the local feature branch
:::

## Pausing and resuming

If you need to switch context mid-PR:

```{r}
usethis::pr_pause()                          # return to main
usethis::pr_resume("my-branch-name")         # pick up where you left off
```

::: {.notes}
This is useful when someone asks you to review their PR while you're mid-work, or when you need to make a quick fix to main.
:::

## Reviewing someone else's PR

```{r}
usethis::pr_fetch(1)    # check out PR #1 locally
```

::: {.fragment}
Run their code, run their tests, leave an informed review on GitHub.

```{r}
usethis::pr_finish()    # clean up when done reviewing
```
:::

## The full picture

```{mermaid}
%%| eval: true
%%| echo: false
flowchart LR
    A([main]) --> PI["pr_init()"] --> B([your branch])
    B --> C[Write & test]
    C --> PP["pr_push()"] --> D([GitHub PR])
    D --> E{Review}
    E -->|Changes needed| C
    E -->|Approved & merged| F([main updated])
    F --> PF["pr_finish()"] --> A

    D --> PFetch["pr_fetch()"] --> G([reviewer's local])
    G -->|review & comment| D

    B --> PA["pr_pause()"] --> A
    A --> PRe["pr_resume()"] --> B

    style A fill:#2d6a4f,color:#fff,stroke:none
    style F fill:#2d6a4f,color:#fff,stroke:none
    style B fill:#1d3557,color:#fff,stroke:none
    style G fill:#1d3557,color:#fff,stroke:none
    style D fill:#457b9d,color:#fff,stroke:none
    style E fill:#e63946,color:#fff,stroke:none
    style C fill:#333,color:#fff,stroke:#666
    style PI fill:#f4a261,color:#000,stroke:none
    style PP fill:#f4a261,color:#000,stroke:none
    style PF fill:#f4a261,color:#000,stroke:none
    style PFetch fill:#f4a261,color:#000,stroke:none
    style PA fill:#f4a261,color:#000,stroke:none
    style PRe fill:#f4a261,color:#000,stroke:none
```

# Continuous Integration

## Why CI?

Your package works on your machine. But does it work on:

::: {.incremental}
- Windows? macOS? Linux?
- R 4.3? R 4.4?
- Someone else's machine with different packages installed?
:::

::: {.fragment}
**Continuous Integration** automatically runs checks on every push - catching problems before they reach collaborators.
:::

## Start a branch

Even infrastructure changes go through a PR - CI validates itself before landing on `main`.

```{r}
usethis::pr_init("add-ci")
```

## Add R CMD check via GitHub Actions

```{r}
usethis::use_github_action("check-standard")
```

::: {.fragment}
Creates `.github/workflows/R-CMD-check.yaml` which runs `R CMD check` on:

- Ubuntu (release R)
- macOS (release R)
- Windows (release R)
- Ubuntu (devel R)
:::

## Add a status badge

The action also adds a badge to your `README.Rmd`:

```{r}
devtools::build_readme()
devtools::check()
```

::: {.fragment}
Then push and open the PR:

```{r}
usethis::pr_push()
```

Watch the Actions tab - CI runs on the PR branch before anything merges to `main`.
:::

## Viewing CI results

After merging with `pr_finish()`, go to the **Actions** tab on your repo.

Each push triggers a workflow run. Green checkmark = passing. Red X = something to fix.

::: {.notes}
Show this live - click into a run and show the output from R CMD check. The output is identical to what you see locally with check(). Getting comfortable reading CI output is a key skill.
:::

# Code Formatting with Air

## What is Air?

[Air](https://posit-dev.github.io/air/) is a fast R code formatter - it automatically reformats your code on save.

**Why it matters for collaboration:**

::: {.incremental}
- Eliminates style debates in code review
- Reduces noisy diffs (no more whitespace-only changes)
- Keeps everyone's code consistent without thinking about it
:::

## Start a branch

```{r}
usethis::pr_init("add-air-formatter")
```

## Set up Air in your project

```{r}
usethis::use_air(vscode = FALSE)
```

Then in RStudio, enable format on save:

- **Tools > Global Options > Code > Formatting** - set formatter to Air
- **Tools > Global Options > Code > Saving** - enable "Format on save"

::: {.fragment}
Air will be downloaded automatically on first save if you haven't installed the CLI.
:::

::: {.notes}
Show the two RStudio settings dialogs. The CLI installation is optional - Air can run without it via the R package, but the CLI is faster.
:::

## Push and merge

```{r}
usethis::pr_push()
```

Review the diff - Air may reformat existing code in the PR. That's fine; it's a one-time cleanup.

```{r}
usethis::pr_finish()
```

# Adding a Function via PR

## Start a branch

```{r}
usethis::pr_init("add-biomass-function")
```

## Write the function

```{r}
usethis::use_r("biomass")
```

```{r}
#' Calculate Biomass Index
#'
#' @param cpue Numeric vector of CPUE values
#' @param area_swept Numeric vector of area swept (km²)
#' @return A numeric vector of biomass index values
#' @export
#' @examples
#' biomass_index(cpue = 10, area_swept = 5)
biomass_index <- function(cpue, area_swept) {
  cpue * area_swept
}
```

## Document and check

```{r}
devtools::document()
devtools::load_all()

biomass_index(10, 5)

devtools::check()
```

Commit when `check()` passes cleanly.

## Push and open a PR

```{r}
usethis::pr_push()
```

::: {.notes}
Walk through the GitHub PR creation UI. Title should match the branch intent. Body should explain *why*, not just *what* - the diff already shows what changed.
:::

# Your Turn

## Exercise

::: {.incremental}
1. Start a branch: `pr_init("add-biomass-function")`
2. Create `R/biomass.R` with the `biomass_index()` function
3. Document with `document()`, check with `check()`
4. Push with `pr_push()` and open a PR on GitHub
5. Review the diff - does the PR description tell the story?
:::

::: {.fragment}
We will use this PR throughout the testing session.
:::

::: {.notes}
Give people 10-15 minutes. Walk around and help anyone who gets stuck on the GitHub PR creation step. The goal is everyone has an open PR by the end of this exercise.
:::
