---
title: "Object-Oriented Programming in R"
format:
  revealjs:
    height: 900
    width: 1600
    theme: [moon, slides.scss]
    highlight-style: a11y
    slide-number: true
    chalkboard: true
    footer: "[{{< fa house >}}](../index.html)"
execute:
  eval: false
  echo: true
---

## Outline

1. What is Object-Oriented Programming and why does R have it?
2. OOP systems in R
3. S3 in depth: classes, attributes, and methods
4. Writing `print`, `summary`, and `plot` methods
5. Creating your own generics

# What is Object-Oriented Programming?

## You already use it

```{r}
#| eval: true
#| output-location: column-fragment
x_norm <- rnorm(5000)
summary(x_norm)
```

\

::: {.fragment}

```{r}
#| eval: true
#| output-location: column-fragment
summary(mtcars)
```

:::

\

::: {.fragment}

```{r}
#| eval: true
#| output-location: column-fragment
mtcars_model <- lm(mpg ~ wt, data = mtcars)
summary(mtcars_model)
```

:::

::: {.notes}
Give participants a moment to think about this. They have been calling summary() for years without thinking about what makes it work. The magic is that the *object* determines what happens, not the caller.
:::

## What makes that possible?

The same function name. Completely different behaviour. No `if/else` in your code.

. . .

Every R object has a **class** attribute:

```{r}
class(x_norm)
#> "numeric"

class(mtcars)
#> "data.frame"

class(mtcars_model)
#> "lm"
```

::: {.fragment}
When you call `summary(x)`, R reads `class(x)` and looks for a function named `summary.<class>`.

This is called **dispatch**: the object's class determines which function runs.
:::

## Generics and methods

::: columns
::: {.column width="50%"}
**A generic** is the user-facing function:

```{r}
summary
```

```
function (object, ...) UseMethod("summary")
```

It does almost nothing itself - it just dispatches.
:::

::: {.column width="50%"}
**A method** is a class-specific implementation:

```{r}
summary.lm
summary.data.frame
summary.default # fallback
```

Named `<generic>.<class>`.
:::
:::

::: {.fragment}
The generic is the consistent interface. The methods are the specialised implementations.
:::

::: {.notes}
Show methods(summary) live if time permits. The list is long - summary has dozens of methods across base R and packages. That's the power of the system: packages can add new methods for existing generics without touching the original code.
:::

## The dispatch chain

When you call `summary(x)` and `class(x)` is `"foo"`:

::: {.incremental}
1. R looks for `summary.foo`
2. If not found, looks for `summary.default`
3. If neither exists - error
:::

::: {.fragment}
This is **single dispatch**: the method is chosen based on the class of the *first argument only*.

```{r}
# What methods exist for a generic?
methods(generic.function = "print")

# What methods exist for a class?
methods(class = "data.frame")
```

:::

# OOP Systems in R

## Four systems, one concept

All four implement the same idea - dispatch based on class - but differ in formality and where methods live.

| System | Dispatch | Formality | Common in |
|--------|----------|-----------|-----------|
| **S3** | Single (first arg) | Minimal | Base R, tidyverse, most packages |
| **S4** | Multiple args | Formal classes, validity checks | Bioconductor |
| **S7** | Multiple args | Formal, successor to S3 + S4 | Some newer packages |
| **R6** | Single | Encapsulated, mutable | Shiny, databases, stateful objects |

::: {.fragment}
S3 covers the vast majority of package development use cases. We focus on S3 today.
:::

::: {.notes}
S4 is worth knowing if you work in Bioconductor or need multiple dispatch. R6 is useful when you need objects that hold mutable state - think a database connection or an HTTP client. S7 is the planned long-term successor to both S3 and S4, combining S3's simplicity with S4's rigour. It is stable but not yet dominant in the ecosystem.
:::

# S3 in Depth

## An S3 object is just a regular R object

...with a `class` attribute attached.

```{r}
x <- c(10, 20, 30)
class(x) # "numeric"

class(x) <- "my_class"
class(x) # "my_class"

# The data is unchanged
x + 1 # still works as a numeric vector
```

::: {.fragment}
That's the whole mechanism. No class declaration. No schema. Just an attribute.
:::

## Giving `cpue()` a class

Currently `cpue()` returns a plain numeric vector. We can tag its output:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  ...
) {
  # ... existing calculation ...

  result <- raw_cpue * gear_factor
  class(result) <- "cpue_result" # tag the output
  result
}
```

::: {.fragment}

```{r}
r <- cpue(c(100, 200, 300), c(10, 20, 15))
class(r) # "cpue_result"
```

:::

::: {.fragment}
Now we can write methods that fire whenever something is done with a `cpue_result`.
:::

## Carrying metadata with attributes

A class tag alone gives us dispatch. **Attributes** let us attach context to the object.

```{r}
attr(result, "method") <- method
attr(result, "gear_factor") <- gear_factor
attr(result, "n_records") <- length(catch)
class(result) <- "cpue_result"
```

::: {.fragment}
Attributes survive assignment and can be read back with `attr()`:

```{r}
attr(r, "method") # "ratio"
attr(r, "n_records") # 3
```

:::

::: {.notes}
Contrast this with a named list approach - attributes keep the numeric vector as the primary data, with metadata attached to it, rather than burying the values inside a list. The object is still a numeric vector that supports arithmetic; it just also carries context.
:::

## The constructor pattern

Setting attributes one by one gets messy. The standard practice is a **constructor function**:

```{r}
new_cpue_result <- function(cpue_values, method, gear_factor, n_records) {
  structure(
    cpue_values, # the data
    method = method, # metadata as named attributes
    gear_factor = gear_factor,
    n_records = n_records,
    class = "cpue_result"
  )
}
```

::: {.fragment}
`structure()` sets all attributes - including `class` - in one call.

Convention: name constructors `new_<classname>`. They are usually internal (`@noRd`).
:::

::: {.fragment}
Then `cpue()` calls the constructor instead of setting attributes directly:

```{r}
result <- new_cpue_result(
  raw_cpue * gear_factor,
  method,
  gear_factor,
  length(catch)
)
```

:::

::: {.notes}
The constructor is the single source of truth for how a cpue_result is built. If you ever need to change what metadata the object carries, there is one place to change it. This matters as the package grows.
:::

# Writing Methods

## The `print` method

When you type an object name at the console, R calls `print()` - this is **autoprinting**.

```{r}
#' @export
print.cpue_result <- function(x, ...) {
  cat("CPUE Result\n")
  cat("Records:     ", attr(x, "n_records"), "\n")
  cat("Method:      ", attr(x, "method"), "\n")
  cat("Gear factor: ", attr(x, "gear_factor"), "\n")
  cat("Values:      ", round(x, 2), "\n")
  invisible(x)
}
```

## Rules for print methods

::: {.incremental}
- Name it `print.<classname>`
- Always include `...` in the signature - the `print` generic requires it
- Return `invisible(x)` so assignment still works: `y <- print(x)`
- Use `cat()` inside, not `print()` - calling `print()` inside `print.foo` causes infinite recursion
- Use `@export` so roxygen2 registers it in NAMESPACE
:::

## The `summary` method

`summary` should compute and return a useful statistical object:

```{r}
#' @export
summary.cpue_result <- function(object, ...) {
  stats <- list(
    method = attr(object, "method"),
    n_records = attr(object, "n_records"),
    gear_factor = attr(object, "gear_factor"),
    mean_cpue = mean(object),
    median_cpue = stats::median(object),
    sd_cpue = stats::sd(object),
    range_cpue = range(object)
  )
  structure(stats, class = "summary.cpue_result")
}
```

::: {.fragment}
It returns an object with its own class - which has its own `print` method.

This is the standard base R pattern: `summary.lm` returns a `summary.lm` object.
:::

## The `plot` method

```{r}
#' @export
plot.cpue_result <- function(x, ...) {
  plot(
    seq_along(x),
    x,
    type = "b",
    xlab = "Record",
    ylab = "CPUE",
    main = paste("CPUE -", attr(x, "method"), "method"),
    ...
  )
}
```

::: {.fragment}
The `...` passes through to `plot.default` - users can customise colour, line type, etc. without us anticipating every option.

You don't have to use base `plot()` - ggplot2 works too.
:::

## Registering methods in NAMESPACE

`@export` on each method tells roxygen2 to write the correct entry:

```{r}
document()
```

```
# NAMESPACE entries
S3method(print,cpue_result)
S3method(summary,cpue_result)
S3method(plot,cpue_result)
```

::: {.fragment}
`S3method()` entries - not plain `export()` entries - are how R finds your methods when the package is installed.

`load_all()` registers them automatically during development, which is why things work in your session even before you call `document()`.
:::

## Testing S3 classes

`expect_s3_class()` checks the class:

```{r}
test_that("cpue() returns a cpue_result object", {
  result <- cpue(c(100, 200), c(10, 20))
  expect_s3_class(result, "cpue_result")
})
```

::: {.fragment}
Test that attributes are set correctly:

```{r}
test_that("cpue_result carries calculation metadata", {
  result <- cpue(c(100, 200, 300), c(10, 20, 15), method = "log")
  expect_equal(attr(result, "method"), "log")
  expect_equal(attr(result, "n_records"), 3)
})
```

:::

::: {.fragment}
Snapshot tests are a natural fit for print methods:

```{r}
test_that("print.cpue_result displays expected output", {
  result <- cpue(c(100, 200, 300), c(10, 20, 15))
  expect_snapshot(print(result))
})
```

:::

## Make a commit

```{r}
document()
check()
```

::: {.fragment}
Commit your class definition, constructor, methods, and tests.
:::

# Creating Your Own Generics

## The motivation

`cpue()` currently takes numeric vectors. But users often have a data frame:

```{r}
# What they have to write now:
cpue(fishing_data$catch, fishing_data$effort)

# What would be nicer:
cpue(fishing_data)
```

::: {.fragment}
We can make `cpue()` work on both. The same function call adapts its behaviour depending on what the user passes in - just like `summary()` does.
:::

## Step 1: Make cpue() a generic

Replace the function body with `UseMethod()`:

```{r}
#' @export
cpue <- function(x, ...) {
  UseMethod("cpue")
}
```

::: {.fragment}
The first argument becomes `x` by convention - it could be a numeric vector or a data frame.

`UseMethod("cpue")` tells R: look at `class(x)` and dispatch to `cpue.<class>`.
:::

## Step 2: The numeric method

Move the existing implementation into `cpue.numeric`:

```{r}
#' @rdname cpue
#' @export
cpue.numeric <- function(
  x,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE),
  ...
) {
  # ... same logic as before ...
  new_cpue_result(raw_cpue * gear_factor, method, gear_factor, length(x))
}
```

::: {.fragment}
All existing calls to `cpue()` with a numeric first argument still work - R dispatches to `cpue.numeric` automatically.
:::

## Step 3: The data frame method

```{r}
#' @rdname cpue
#' @export
cpue.data.frame <- function(
  x,
  catch_col = "catch",
  effort_col = "effort",
  ...
) {
  if (!catch_col %in% names(x)) {
    stop("Column '", catch_col, "' not found.", call. = FALSE)
  }
  if (!effort_col %in% names(x)) {
    stop("Column '", effort_col, "' not found.", call. = FALSE)
  }

  cpue(x[[catch_col]], effort = x[[effort_col]], ...)
}
```

::: {.fragment}
The data frame method extracts columns and calls `cpue()` again with numeric vectors - which dispatches to `cpue.numeric`.

This **delegation pattern** - one method calling the generic with different input - is common and effective. The calculation logic lives in exactly one place.
:::

## Step 4: The default method

```{r}
#' @rdname cpue
#' @export
cpue.default <- function(x, ...) {
  stop("Unsupported input type for cpue(): ", class(x), call. = FALSE)
}
```

::: {.fragment}
Without a default, passing an unsupported type gives a cryptic error: "no applicable method for 'cpue' applied to an object of class ...".

With a default, users get a message that names the function and the problem.
:::

## Using `@rdname` to group docs

All three methods share one help page:

```{r}
#' Calculate CPUE
#' @param x A numeric vector of catch, or a data frame.
#' @param ... Additional arguments passed to methods.
#' @export
cpue <- function(x, ...) UseMethod("cpue")

#' @rdname cpue
#' @export
cpue.numeric <- function(x, effort, ...) {
  ...
}

#' @rdname cpue
#' @export
cpue.data.frame <- function(x, catch_col = "catch", ...) {
  ...
}
```

::: {.fragment}
`@rdname cpue` tells roxygen2 to add documentation to the `cpue` help page rather than create a new one.
:::

## It works

```{r}
load_all()

# Vector - dispatches to cpue.numeric
cpue(c(100, 200, 300), c(10, 20, 15))

# Data frame - dispatches to cpue.data.frame
fishing_data <- data.frame(catch = c(100, 200), effort = c(10, 20))
cpue(fishing_data)

# Custom column names
trawl_data <- data.frame(kg = c(50, 100), hours = c(5, 10))
cpue(trawl_data, catch_col = "kg", effort_col = "hours")
```

## One thing to watch for

The first argument was renamed from `catch` to `x`.

Any call using the named argument `catch = ` will break:

```{r}
cpue(catch = 100, effort = 10) # Error: unused argument (catch = 100)
cpue(100, 10) # Fine - positional arguments still work
```

::: {.fragment}
Existing tests that use `catch =` will need updating. This is a **breaking change** - something to note in your changelog.
:::

## Make a commit

```{r}
document()
check()
snapshot_accept() # print format changed - accept updated snapshots
```

::: {.fragment}
Commit the generic, methods, updated documentation, and tests.
:::

# When to Use OOP

## S3 is worth it when...

::: {.incremental}
- Your function returns complex results that benefit from custom `print`/`summary`/`plot`
- You want the same function name to work on multiple input types (vectors, data frames, model objects)
- You want other packages to be able to extend your work by writing new methods
:::

## S3 is overkill when...

::: {.incremental}
- A named list or data frame communicates the result clearly on its own
- There is only one type of input and one way to display the output
- The function returns a scalar or vector with obvious meaning
:::

::: {.fragment}
Start simple. Reach for S3 when a plain return value creates friction for users.
:::

## Recap

::: {.incremental}
- S3 objects are regular R objects with a `class` attribute
- Generics dispatch to `generic.class` based on the class of the first argument
- Constructors (`new_<classname>`) centralise object creation with `structure()`
- `print`, `summary`, and `plot` methods make your classes pleasant to work with
- Create your own generics with `UseMethod()` for consistent interfaces across input types
- `@export` on each method registers it in NAMESPACE via `S3method()`
- Test classes with `expect_s3_class()` and snapshot tests for print output
:::
