---
title: "Testing with testthat"
subtitle: "Part 2"
format:
  revealjs:
    height: 900
    width: 1600
    theme: [moon, slides.scss]
    highlight-style: a11y
    slide-number: true
    chalkboard: true
    footer: "[{{< fa house >}}](../index.html)"
execute:
  eval: false
  echo: true
---

## Outline

1. Updating snapshots
2. Writing clean tests with `withr`
3. Test coverage with `covr`

# Updating Snapshots

## Where we left off

In "Function Design Best Practices" we added input validation with better error messages.

::: {.fragment}
Our snapshot tests captured the *old* error output - they will now fail.
:::

::: {.fragment}
```{r}
test()
```
:::

::: {.notes}
Give participants a moment to run test() so they can see the failure. The failure message will say something like "Snapshot of `cpue("not a number", 10)` has changed". This is expected - walk them through why.
:::

## What a snapshot failure looks like

```
── Failure (test-cpue.R:14:3): cpue error message is informative ──────────────
Snapshot of `cpue("not a number", 10)` has changed:
old[3] == "  non-numeric argument to binary operator"
new[3] == "  catch must be numeric, not character"
```

::: {.fragment}
The test is doing its job: it detected that your output changed.
:::

::: {.fragment}
Now you decide: is this the output you *intended*?
:::

## Reviewing snapshot changes

```{r}
snapshot_review()   # interactive diff viewer
```

::: {.fragment}
If the new output is correct:

```{r}
snapshot_accept()   # accept all pending changes
```
:::

::: {.notes}
snapshot_review() opens an interactive Shiny app showing old vs. new output side by side. Walk through it live. snapshot_accept() writes the new output to the _snaps/ files. Remind participants to commit the updated snapshots.
:::

## The snapshot workflow

::: {.incremental}
1. Change something (a function, an error message, a print method)
2. Run `test()` - snapshot tests fail
3. Run `snapshot_review()` - inspect the diff
4. Accept if the change is intentional, fix the code if it is not
5. Commit the updated `.md` files in `tests/testthat/_snaps/`
:::

::: {.fragment}
::: {.notebox}
This cycle will repeat throughout the course. That is normal and expected.
:::
:::

# Writing Clean Tests

## Tests and global state

A test that changes global state can affect tests that run after it, 
_AND_ it can affect the user’s session if they run tests interactively.

::: {.fragment}
**Global state includes:**

- Options (`getOption()`)
- Environment variables
- Working directory
- Random seed
- Files on disk
:::

::: {.fragment}
The fix: make every change *local* to the test that needs it.
:::

::: {.notes}
global state is anything that persists across function calls and affects how code behaves - shared mutable context that outlives the call that changed it.
:::

## withr

The `withr` package provides helpers to make temporary changes that are automatically reversed.

::: {.fragment}

```{r}
use_package("withr", type = "Suggests")
```

:::

::: {.fragment .question}
Aside: Why do we set `type = "Suggests"` for withr?
:::

## Two families of `withr_` helpers:

| Prefix | Scope | Use for |
|--------|-------|---------|
| `local_*` | Runs until the current environment exits | Inside `test_that()` |
| `with_*` | Runs for one expression | Wrapping a single call |

examples:

::: {.columns}
::: {.column}

``` r
getOption("digits")
#> [1] 7
pi
#> [1] 3.141593

withr::with_options(
  list(digits = 2),
  print(pi)
)
#> [1] 3.1

getOption("digits")
#> [1] 7
```

:::
::: {.column}

```{r}
Sys.time()
#> [1] "2026-02-23 12:24:34 PST"

withr::with_timezone(
  "Europe/Paris", 
  print(Sys.time())
)
#> [1] "2026-02-23 21:25:36 CET"

withr::with_timezone(
  "America/New_York", 
  print(Sys.time())
)
#> [1] "2026-02-23 15:26:13 EST"

Sys.time()
#> [1] "2026-02-23 12:26:39 PST"
```

:::
:::


## Recall: verbose via a package option

`cpue()` reads verbosity from an option:

```{r}
cpue <- function(
  catch,
  effort,
  gear_factor = 1,
  method = c("ratio", "log"),
  verbose = getOption("fishr.verbose", FALSE)
) { ... }
```

::: {.fragment}
Tests that set `options(fishr.verbose = TRUE)` must restore the original value afterward - or they will leak into other tests.
:::

## Testing options with `local_options`

```{r}
test_that("cpue uses verbosity when option set to TRUE", {
  withr::local_options(fishr.verbose = TRUE) # reset when this block exits

  expect_snapshot(cpue(100, 10))
})

test_that("cpue is not verbose when option set to FALSE", {
  withr::local_options(fishr.verbose = FALSE) # reset when this block exits

  expect_silent(cpue(100, 10))
})
```

::: {.notes}
Emphasise that without local_options, the first test would leave fishr.verbose = TRUE in the session, which would cause the second test to fail even though the logic is correct. This is a classic test pollution bug.
:::

## `with_options` for a single assertion

Provide options in a list

```{r}
test_that("cpue verbosity falls back to FALSE when not set", {
  withr::with_options(
    list(fishr.verbose = NULL), # reset as soon as this expression finishes
    expect_no_message(cpue(100, 10))
  )
})
```

# Your Turn

## Exercise

Add the same `verbose` argument to `biomass_index()` and write clean tests for it.

::: {.fragment}
1. Update `R/biomass.R` to read from `getOption("fishr.verbose", FALSE)`
2. Open the test file: `use_test("biomass")`
3. Write tests using `withr::local_options()`
4. Run `test()` and accept any new snapshots
:::

::: {.fragment}
::: {.question}
How would you test that the verbose message is correct?
:::
:::

## Solution

```{r}
test_that("biomass_index uses verbosity when set as an option", {
  withr::local_options(fishr.verbose = TRUE)

  expect_snapshot(biomass_index(cpue = 5, area_swept = 100))
})

test_that("biomass_index verbosity falls back to FALSE when not set", {
  withr::local_options(fishr.verbose = NULL)

  expect_no_message(biomass_index(cpue = 5, area_swept = 100))
})
```

# More withr Helpers

## Common helpers

| Helper | Resets |
|--------|--------|
| `local_options()` / `with_options()` | R options |
| `local_envvar()` / `with_envvar()` | Environment variables |
| `local_dir()` / `with_dir()` | Working directory |
| `local_tempfile()` / `with_tempfile()` | Temp file (deleted after) |
| `local_tempdir()` / `with_tempdir()` | Temp directory (deleted after) |
| `local_seed()` / `with_seed()` | Random seed |

## Example: testing file output

```{r}
test_that("results can be written to a file", {
  outfile <- withr::local_tempfile(fileext = ".csv")

  write_cpue_results(cpue(100, 10), file = outfile)

  expect_true(file.exists(outfile))
  expect_equal(nrow(read.csv(outfile)), 1)
})
# outfile is deleted automatically after the test
```

## Quiz

::: {.question}
Where have we already altered the global state in our test helpers?
:::

::: {.fragment}
`tests/testthat/helper.R` calls `set.seed()` directly - this sets the seed for the entire session.

```{r}
# Before
generate_fishing_data <- function(n = 10) {
  set.seed(67)
  data.frame(...)
}
```

:::

::: {.fragment}

```{r}
# After
generate_fishing_data <- function(n = 10) {
  withr::local_seed(67)
  data.frame(...)
}
```

:::

::: {.notes}
local_seed() sets the seed, runs the body of the function, then restores the previous seed state. This means calling generate_fishing_data() no longer has any effect on the random number stream outside of the function.
:::

<!-- # Test Coverage with covr

## Why measure coverage?

Coverage tracks which lines of your code are *executed* when your tests run.

::: {.incremental}
- Reveals untested code paths - the gaps you didn't notice
- Gives you an objective starting point when inheriting a package
- CI can report coverage changes on every PR
:::

::: {.fragment}
Coverage tells you where tests are *absent*. It does not tell you whether the tests you have are any good.
:::

## Set up covr

```{r}
use_package("covr", type = "Suggests")
```

## Run package coverage

```{r}
library(covr)

cov <- package_coverage()
cov
```

::: {.fragment}
Prints a per-file coverage percentage:

```
fishr Coverage: 87.50%
R/biomass.R: 75.00%
R/cpue.R:    93.75%
```
:::

::: {.fragment}
For a single overall number:

```{r}
percent_coverage(cov)
```
:::

::: {.notes}
package_coverage() loads the package, runs all your tests, and records which lines were hit. It takes the same time as running test(). The output already tells a useful story - which files have the most gaps.
:::

## Finding untested code

```{r}
zero_coverage(cov)
```

::: {.fragment}
Returns a data frame of every line with zero hits. In RStudio, these appear as clickable markers - you can jump straight to the uncovered line.
:::


## Visual report

```{r}
report(cov)
```

::: {.fragment}
Opens an HTML page showing each file with lines colour-coded:

- Green: covered
- Red: not covered
:::

::: {.notes}
Show this live. The visual report makes it immediately obvious which branches are untested - for example, an error-handling path that is never triggered by tests.
:::

## Active file coverage

```{r}
devtools::test_coverage_active_file()
```

::: {.fragment}
Runs coverage only for the file you are editing - much faster than the full package.

Consider binding this to a keyboard shortcut: `Ctrl/Cmd + R`
:::

## Excluding lines

Some lines are not worth testing - interactive-only code paths, print methods, and so on.

```{r}
if (interactive()) {
  # nocov start
  browse_results(x)
} # nocov end
```

::: {.fragment}
Or exclude a single line:

```{r}
stop("Column must be numeric") # nocov
```

:::

## Coverage in CI

Add a GitHub Actions workflow that reports coverage on every push:

```{r}
usethis::use_github_action("test-coverage")
```

::: {.fragment}
By default this uploads results to [Codecov](https://about.codecov.io/), which tracks coverage over time and comments on pull requests with coverage diffs.

You will need to add a `CODECOV_TOKEN` repository secret - see the session notes for setup details.
:::

::: {.notes}
If participants do not want a Codecov account, they can edit the generated workflow to remove the codecov/codecov-action upload step. Coverage will still be printed to the Actions log on every run.
:::

## A word on coverage targets

::: {.incremental}
- High coverage is useful as a guide for finding gaps
- Chasing 100% is rarely worthwhile
- Coverage tells you which lines were *executed*, not whether the tests verify correct behaviour
- A test that runs code without assertions inflates the number without adding value
:::

::: {.fragment}
Focus on covering the important logic paths and edge cases - not the number.
::: -->
